---
title: "POD Application HDBSCAN CL List Template Cluster Analysis"
subtitle: Cluster using the first two PCs from PCA, centered and unscaled
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
# Set working directory as project root directory 
knitr::opts_knit$set(root.dir = ".")
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Import libraries
library(readr)
library(tidyverse)
library(cluster)
library(clusterCrit)
library(uwot)
library(proxy)
library(kableExtra)
library(forcats)
library(lubridate)
library(survival)
library(stringr)
library(scales)
library(webshot2)

# Import data 
cluster_assignments <- read_csv("data/cohorts/01_POD/HDBSCAN_clusters/pod_hdbscan_cl_list_clusters_2pcs_centered_unscaled_100minPts.csv")
all_pod_study_cohort_features <- read_csv("data/cohorts/final_cohort_features.csv")
pod_medications <- read_csv("data/derived/pod_medications_cleaned.csv")
embeddings <- read_csv("data/embeddings/01_POD/pod_clinical_longformer_list_template_embeddings.csv")
rass_scores <- read_csv("data/derived/01_POD/pod_pos_rass_scores_cleaned.csv")
```

# Data Cleaning

```{r data cleaning, echo=TRUE, message=FALSE, warning=FALSE}
# Add cluster assignment column to pod_study_cohort_features
all_pod_study_cohort_features <- left_join(all_pod_study_cohort_features, cluster_assignments, by = "subject_id")

# Filter out HDBSCAN noise observations from pod_study_cohort_features 
noise_observations <- all_pod_study_cohort_features %>% filter(cluster == 0)
pod_study_cohort_features <- all_pod_study_cohort_features %>% filter(cluster != 0)

# Add column to pod_study_cohort_features for days from first surgery to first positive CAM-ICU result
pod_study_cohort_features <- pod_study_cohort_features %>% mutate(first_surgery_to_pod_pos_days = as.numeric(difftime(first_cam_positive_time, first_surgery_date, units = "days")))

# Add binary indicator column for in-hospital mortality to pod_study_cohort_features 
pod_study_cohort_features <- pod_study_cohort_features %>%
  mutate(mortality = case_when(is.na(deathtime) == TRUE ~ "No",
         is.na(deathtime) == FALSE ~ "Yes"))

# Add column for delirium endtime to ponv_study_cohort_features (defined as first_cam_positive_time + pod_resolution_days)
pod_study_cohort_features <- pod_study_cohort_features %>%
  mutate(
    delirium_endtime = ymd_hms(first_cam_positive_time) + ddays(pod_resolution_days)
  )

# Filter pod_medications to subjects in pod_study_cohort_features
pod_medications_filtered <- pod_medications %>% filter(subject_id %in% pod_study_cohort_features$subject_id)

# Add columns to pod_medications_filtered
pod_medications_filtered <- left_join(pod_medications_filtered, pod_study_cohort_features %>% select(subject_id, first_surgery_date, first_cam_positive_time, delirium_endtime, pod_resolution_days, los_days, cluster), by = "subject_id")

# Rename Rivastigmine Tartrate as Rivastigmine
pod_medications_filtered <- pod_medications_filtered %>% mutate(medication = case_when(medication == "Rivastigmine Tartrate" ~ "Rivastigmine", TRUE ~ medication))

# Rename routes to uppercase 
pod_medications_filtered <- pod_medications_filtered %>%
  mutate(
    route = str_to_sentence(route),
    route = str_replace(route, "^Iv\\b", "IV")
  )

# Filter rass_scores to subjects in pod_study_cohort_features
rass_scores_filtered <- rass_scores %>% filter(subject_id %in% pod_study_cohort_features$subject_id)

# Add columns to rass_scores_filtered 
rass_scores_filtered <- left_join(rass_scores_filtered, pod_study_cohort_features %>% select(subject_id, first_surgery_date, first_cam_positive_time, delirium_endtime, cluster), by = "subject_id")

# Add column for total POD medications administered to pod_study_cohort_features 
pod_med_admin <- pod_medications_filtered %>%
  count(subject_id, name = "total_pod_med_admin")

pod_study_cohort_features <- pod_study_cohort_features %>%
  left_join(pod_med_admin, by = "subject_id") %>%
  mutate(
    total_pod_med_admin = coalesce(total_pod_med_admin, 0L))
```

# Colors and Themes 

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Colorblind friendly cluster colors 
cluster_colors <- c(
  "1" = "#56B4E9",
  "2" = "#E69F00",  
  "3" = "#009E73",  
  "4" = "#0072B2",  
  "5" = "#D55E00",  
  "6" = "#CC79A7",
  "7" = "#F0E442",
  "8" = "#999999")

# Nature themed ggplot
theme_nature <- theme_minimal(base_size = 11, base_family = "") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.3),
    axis.ticks = element_line(color = "black", size = 0.3),
    axis.text  = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
    legend.position = "right",
    legend.text  = element_text(size = 11),
    legend.title = element_text(size = 11)
  )
```

# 2D PCA Cluster Visualization

```{r, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE}
# 2D PCA visualization of the clusters

# Filter embeddings to subjects assigned to a cluster
cl_list_emb_filtered <- embeddings %>% filter(subject_id %in% pod_study_cohort_features$subject_id)

# PCA
pca_result <- prcomp(cl_list_emb_filtered %>% select(-subject_id), center = TRUE, scale. = FALSE)
pca_data <- pca_result$x[, 1:2]
pca_data <- data.frame(pca_data)

# Add subject_id and cluster assignment back to pca_data
pca_clean <- cbind(pod_study_cohort_features %>% select(subject_id, cluster), pca_data)

# 2D PCA for cluster visualization
plot_df <- data.frame(
    PC1 = pca_clean$PC1,
    PC2 = pca_clean$PC2,
    Cluster = factor(pca_clean$cluster)  # factor for discrete colors
  )

# PCA Plot
pca_plot <- ggplot(plot_df, aes(x = PC1, y = PC2, color = Cluster)) +
    geom_point(size = 2, alpha = 0.8) +
    scale_color_manual(values = cluster_colors) +
    labs(
      #title = "2D PCA visualization of the clusters",
      #subtitle = "Only center embeddings and keep the first two PCs",
      x = "PC 1",
      y = "PC 2"
    ) + theme_nature

pca_plot

ggsave(
  "figures/main/01_POD/pod_2d_pca_plot.png",
  pca_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

# 2D UMAP Cluster Visualization

```{r, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(uwot)

# Filter embeddings to subjects assigned to a cluster
cl_list_emb_filtered <- embeddings %>%
  filter(subject_id %in% pod_study_cohort_features$subject_id) %>%
  arrange(subject_id)

# Make sure cluster labels are aligned by subject_id
cluster_df <- pod_study_cohort_features %>%
  select(subject_id, cluster) %>%
  arrange(subject_id)

# UMAP input matrix
embedding_mat <- cl_list_emb_filtered %>%
  select(-subject_id) %>%
  as.matrix()

# 2D UMAP
set.seed(42)

umap_result <- uwot::umap(
  embedding_mat,
  n_neighbors = 50,
  min_dist = 0.3,
  n_components = 2,
  metric = "cosine",
  scale = FALSE,
  init = "spectral",
  n_threads = parallel::detectCores() - 1,
  verbose = FALSE
)

# Add subject_id and cluster assignment
umap_data <- as.data.frame(umap_result) %>%
  rename(UMAP1 = V1, UMAP2 = V2) %>%
  mutate(subject_id = cl_list_emb_filtered$subject_id) %>%
  left_join(cluster_df, by = "subject_id") %>%
  mutate(Cluster = factor(cluster))

# UMAP plot
umap_plot <- ggplot(umap_data, aes(x = UMAP1, y = UMAP2, color = Cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = cluster_colors) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  theme_nature

umap_plot

ggsave(
  "figures/main/01_POD/pod_2d_umap_plot.png",
  umap_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

# Cluster Summary Table

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Data for KM estimation
surv_data <- pod_study_cohort_features %>%
  mutate(
    cluster = factor(cluster, labels = sort(unique(cluster))),
    time  = pod_resolution_days,
    event = as.integer(!censored)   # 1 = POD resolved, 0 = censored
  )

# Fit KM by cluster
km_fit <- survfit(Surv(time, event) ~ cluster, data = surv_data)

# Extract KM-estimated median and 95% CI
km_medians <- summary(km_fit)$table %>%
  as.data.frame() %>%
  tibble::rownames_to_column("cluster") %>%
  mutate(
    cluster = sub("^cluster=", "", cluster),
    km_median_pod_resolution = paste0(
      sprintf("%.2f", median),
      " (",
      sprintf("%.2f, %.2f", `0.95LCL`, `0.95UCL`),
      ")"
    )
  ) %>%
  select(cluster, km_median_pod_resolution)

# Extract KM-estimated % resolved by day 7
km_day7 <- summary(km_fit, times = 7) %>%
  with(
    data.frame(
      cluster = sub("^cluster=", "", strata),
      km_percent_resolved_day7 = paste0(
        sprintf("%.1f", 100 * (1 - surv)),
        "%")))

# Create cluster summary 
cluster_summary <- pod_study_cohort_features %>%
  mutate(
    cluster = factor(cluster, labels = sort(unique(cluster))),
    pod_medication_use = factor(pod_medication_use, levels = c("No", "Yes"))
  ) %>%
  group_by(cluster) %>%
  summarise(
    count = n(),

    pod_medication_use_binary = ifelse(
      any(pod_medication_use == "Yes"),
      "Yes",
      "No"
    ),

    total_pod_med_admin = paste0(
      sprintf("%.0f", median(total_pod_med_admin, na.rm = TRUE)),
      " (",
      sprintf("%.0f, %.0f",
              min(total_pod_med_admin, na.rm = TRUE),
              max(total_pod_med_admin, na.rm = TRUE)),
      ")"
    ),

    age_mean_sd = paste0(
      sprintf("%.1f", mean(age, na.rm = TRUE)),
      " (",
      sprintf("%.1f", sd(age, na.rm = TRUE)),
      ")"
    ),

    percent_male = paste0(sprintf(
      "%.1f",
      100 * mean(gender == "Male", na.rm = TRUE)
    ), "%"),

    percent_ventilator = paste0(sprintf(
      "%.1f",
      100 * mean(ventilator_use == "Yes", na.rm = TRUE)
    ), "%"),

    .groups = "drop"
  ) %>%
  left_join(km_medians, by = "cluster") %>%
left_join(km_day7, by = "cluster")

colnames(cluster_summary) <- c(
  "Cluster",
  "Count",
  "POD Medication Use",
  "Median POD Med Admin (Min, Max)",
  "Mean Age (SD)",
  "% Male",
  "% Ventilator Use",
  "Median POD Resolution Days (95% CI)",
  "% Resolved by Day 7"
)

# Ensure named vector for colors
cluster_color_map <- cluster_colors
names(cluster_color_map) <- c(1,2,3,4,5,6)

cluster_table_kable <- cluster_summary %>%
  kable(
    format = "html",
    escape = FALSE,
    align = "lcccccccc"
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = NULL,
    font_size = 14
  ) %>%

  # Fix column widths (this enables wrapping)
  column_spec(1, width = "1.5cm") %>%   # Cluster
  column_spec(2, width = "1.5cm") %>%   # Count
  column_spec(3, width = "3.5cm") %>%   # POD Medication Use
  column_spec(4, width = "3.5cm") %>%   # Median POD Med Admin
  column_spec(5, width = "2.6cm") %>%   # Age
  column_spec(6, width = "2.4cm") %>%   # Gender
  column_spec(7, width = "2.8cm") %>%   # Ventilator use 
  column_spec(8, width = "4.0cm") %>%   # Median POD resolution time
  column_spec(9, width = "3.5cm") %>%   # % Resolved by Day 7
  column_spec(
    1,
    bold = TRUE,
    background = cluster_color_map[cluster_summary$Cluster],
    color = "black"
  )

cluster_table_kable

# Save table as a PNG
save_kable(
  cluster_table_kable,
  file = "figures/main/01_POD/pod_cluster_summary_table_for_figure.png",
  zoom = 2
)
```

# Cluster Phenotype Table

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cluster_summary <- data.frame(
  Cluster = c(1, 2, 3, 4, 5, 6),
  Phenotype = c("Brief delirium without pharmacologic intervention",
                "Prolonged delirium with intensive sedative use and sustained mechanical ventilation",
                "Intermediate-resolution delirium with sedative-heavy management and frequent mechanical ventilation",
         "Intermediate-resolution delirium with balanced pharmacologic management, favoring sedatives",
         "Intermediate-resolution delirium with balanced pharmacologic management, favoring antipsychotics",
                "Intermediate-resolution delirium with targeted pharmacologic management (predominantly antipsychotics)")
)
  
# Ensure named vector for colors
cluster_color_map <- cluster_colors
names(cluster_color_map) <- levels(cluster_summary$Cluster)

cluster_table_kable <- cluster_summary %>%
  kable(
    format = "html",
    escape = FALSE,
    align = "lc"
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = NULL,
    font_size = 14
  ) %>%

  # Fix column widths (this enables wrapping)
  column_spec(1, width = "1.5cm") %>%   # Cluster
  column_spec(2, width = "10.5cm") %>%   # Phenotype 
  column_spec(
    1,
    bold = TRUE,
    background = cluster_color_map[cluster_summary$Cluster],
    color = "black"
  )

cluster_table_kable

# Save table as a PNG
save_kable(
  cluster_table_kable,
  file = "figures/main/01_POD/pod_cluster_phenotype_table_for_figure.png",
  zoom = 2
)
```


# POD Medications

## Medication Administration Percentages Table

```{r, echo=FALSE, message=FALSE}
medication_pct_table <- pod_medications_filtered %>%
  group_by(cluster, medication) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(
    percentage = round(100 * n / sum(n), 2)
  ) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(
    names_from = cluster,
    values_from = percentage,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x), -medication)

kable(medication_pct_table, caption = "POD Administration Percentages by Cluster")
```

## Radar Plot for POD Medication Administration Percentages

```{r, echo=FALSE, message=FALSE, warning=FALSE}
medication_order <- c("Dexmedetomidine", "Quetiapine", "Haloperidol", "Olanzapine", "Risperidone", "Chlorpromazine", "Rivastigmine")

medication_counts <- pod_medications_filtered %>% 
  mutate(medication = factor(medication, levels = medication_order)) %>%
  group_by(cluster, medication) %>%
  summarise(n_admin = n(), .groups = "drop_last") %>%
  mutate(percentage = round(100 * n_admin / sum(n_admin),2)) %>%
  arrange(cluster, desc(percentage))

# Step 1: Wide format (clusters as rows, meds as columns)
radar_data <- medication_counts %>%
  select(cluster, medication, percentage) %>%
  pivot_wider(names_from = medication, values_from = percentage, values_fill = 0)

radar_data <- radar_data %>%
  select(cluster, all_of(medication_order))

# Step 2: Add required max/min rows for fmsb
radar_scaled <- rbind(
  rep(100, ncol(radar_data)-1),
  rep(0, ncol(radar_data)-1),
  radar_data[,-1]
)
rownames(radar_scaled) <- c("max", "min", paste0("Cluster ", radar_data$cluster))

# Extract colors in cluster order
radar_cluster_ids <- as.character(radar_data$cluster)
radar_colors <- cluster_colors[radar_cluster_ids]

# Save as a png image
png(
  filename = "figures/main/01_POD/pod_medication_admin_pct_radar.png",
  width  = 7.0,      # inches
  height = 4.5,      # inches
  units  = "in",
  res    = 300,      # DPI for journals
  bg     = "white"
)

# Adjust the margins 
par(mar = c(0, 0, 0, 2.8)) # bottom, left, top, right

fmsb::radarchart(radar_scaled,
                 axistype = 0,
                 pcol = radar_colors,
                 plwd = 0.5, # line width
                 plty = 1,
                 cglcol = "grey",
                 cglty = 1,
                 axislabcol = "grey20",
                 caxislabels = c(0, 20, 40, 60, 80, 100),
                 cglwd = 0.8,
                 vlcex = 0.9, # medication label sizes
                 cex.main = 0.8)

legend("topright",
       legend = paste0("Cluster ", radar_cluster_ids),
       col = radar_colors,
       lty = 1, lwd = 1.4, cex = 0.8,
       inset = c(-0.05, 0.05),
       xpd = TRUE)

# Close device
dev.off()
```

## Administration Modes

```{r, echo=FALSE, message=FALSE}
route_pct_table <- pod_medications_filtered %>%
  group_by(cluster, route) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(
    percentage = round(100 * n / sum(n), 2)
  ) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(
    names_from = cluster,
    values_from = percentage,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x), -route)

kable(route_pct_table, caption = "POD Administration Routes by Cluster")
```

## Radar Plot for Administration Modes

```{r, echo=FALSE, message=FALSE, warning=FALSE}
route_counts <- pod_medications_filtered %>% 
  group_by(cluster, route) %>%
  summarise(n_admin = n(), .groups = "drop_last") %>%
  mutate(percentage = round(100 * n_admin / sum(n_admin),2)) %>%
  arrange(cluster, desc(percentage))

# Step 1: Wide format (clusters as rows, meds as columns)
radar_data <- route_counts %>%
  select(cluster, route, percentage) %>%
  pivot_wider(names_from = route, values_from = percentage, values_fill = 0)

# Step 2: Add required max/min rows for fmsb
radar_scaled <- rbind(
  rep(100, ncol(radar_data)-1),
  rep(0, ncol(radar_data)-1),
  radar_data[,-1]
)
rownames(radar_scaled) <- c("max", "min", paste0("Cluster ", radar_data$cluster))

# Extract colors in cluster order
radar_cluster_ids <- as.character(radar_data$cluster)
radar_colors <- cluster_colors[radar_cluster_ids]

# Save as a png image
png(
  filename = "figures/main/01_POD/pod_medication_routes_radar.png",
  width  = 7.0,      # inches
  height = 4.5,      # inches
  units  = "in",
  res    = 300,      # DPI for journals
  bg     = "white"
)

# Adjust the margins 
par(mar = c(0, 0, 0, 2.8)) # bottom, left, top, right

fmsb::radarchart(radar_scaled,
                 axistype = 0,
                 pcol = radar_colors,
                 plwd = 0.5, # line width
                 plty = 1,
                 cglcol = "grey",
                 cglty = 1,
                 axislabcol = "grey20",
                 caxislabels = c(0, 20, 40, 60, 80, 100),
                 cglwd = 0.8,
                 vlcex = 0.9, # route label sizes
                 cex.main = 0.8)

legend("topright",
       legend = paste0("Cluster ", radar_cluster_ids),
       col = radar_colors,
       lty = 1, lwd = 1.4, cex = 0.8,
       inset = c(-0.05, 0.05),
       xpd = TRUE)

# Close device
dev.off()
```

## Medication Type Switches (Mean switches per cluster)

```{r, echo=FALSE, message=FALSE}
# Order medications and identify switches within each subject
pod_switches <- pod_medications_filtered %>%
  arrange(subject_id, starttime) %>%
  group_by(subject_id) %>%
  mutate(
    prev_medication = lag(medication),
    is_switch = !is.na(prev_medication) & medication != prev_medication
  ) %>%
  ungroup()

# Summarize switches per subject
switch_summary <- pod_switches %>%
  group_by(subject_id, cluster) %>%
  summarise(
    n_switches = sum(is_switch),
    .groups = "drop"
  ) %>%
  group_by(cluster) %>%
  summarise(
    mean_switches = mean(n_switches),
    sd_switches   = sd(n_switches),
    n_subjects    = n(),
    .groups = "drop"
  )

# Bar plot for mean number of switches with error bars
ggplot(
  switch_summary,
  aes(x = factor(cluster), y = mean_switches, fill = factor(cluster))
) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = mean_switches - sd_switches / sqrt(n_subjects),
      ymax = mean_switches + sd_switches / sqrt(n_subjects)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_switches + sd_switches / sqrt(n_subjects),
    label = sprintf("%.1f", mean_switches)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Mean medication switches per patient"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))+
  theme_nature
```

## Medication Type Switches (Percentage per cluster)

```{r, echo=FALSE, message=FALSE}
# Order medications and identify switches within each subject
pod_switches <- pod_medications_filtered %>%
  arrange(subject_id, starttime) %>%
  group_by(subject_id) %>%
  mutate(
    prev_medication = lag(medication),
    is_switch = !is.na(prev_medication) & medication != prev_medication
  ) %>%
  ungroup()

# Subject-level switch indicator
switch_subject <- pod_switches %>%
  group_by(subject_id, cluster) %>%
  summarise(
    any_switch = any(is_switch),
    .groups = "drop"
  )

switch_percent <- switch_subject %>%
  group_by(cluster) %>%
  summarise(
    n_switched        = sum(any_switch),
    n_subjects        = n(),
    percent_switched  = 100 * n_switched / n_subjects,
    .groups = "drop"
  )

switch_pct_plot <- ggplot(
  switch_percent,
  aes(x = factor(cluster), y = percent_switched, fill = factor(cluster))
) +
  geom_col(width = 0.7) +
  geom_text(
     aes(
    label = paste0(
      sprintf("%.1f%%", percent_switched),
      "\n(n=", n_switched, ")"
    )),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Patients with ≥1 medication switch (%)"
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_nature

switch_pct_plot

ggsave(
  "figures/main/01_POD/pod_medication_switch_percentages_by_cluster_bar_plot.png",
  switch_pct_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## Medication Type Switches (Most common switches)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pod_switches <- pod_medications_filtered %>%
  arrange(subject_id, starttime) %>%
  group_by(subject_id) %>%
  mutate(
    prev_medication = lag(medication),
    is_switch = !is.na(prev_medication) & medication != prev_medication
  ) %>%
  ungroup()

# Extract the most common switches 
common_switches <- pod_switches %>%
  filter(is_switch) %>%
  mutate(
    switch = paste(prev_medication, "→", medication)
  ) %>%
  count(cluster, switch, sort = TRUE)

top_medication_switches_overall <- common_switches %>%
  group_by(switch) %>%
  summarise(
    total_n = sum(n),
    .groups = "drop"
  ) %>%
  arrange(desc(total_n)) %>%
  slice_head(n = 10)

top_switches_by_cluster <- common_switches %>%
  group_by(cluster) %>%
  slice_max(n, n = 3) %>%
  ungroup()

kable(top_switches_by_cluster, caption = "Top medication switches by cluster")
```

## Medication Type Switches (Heatmap)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
med_class_map <- tibble(
  medication = c(
    "Dexmedetomidine",
    "Quetiapine",
    "Haloperidol",
    "Olanzapine",
    "Risperidone",
    "Chlorpromazine",
    "Rivastigmine"
  ),
  med_class = c(
    "Sedative",          # Dexmedetomidine
    "Antipsychotic",
    "Antipsychotic",
    "Antipsychotic",
    "Antipsychotic",
    "Antipsychotic",
    "Other"              # Rivastigmine (exclude later)
  )
)

pod_switches_class <- pod_switches %>%
  filter(is_switch) %>%
  left_join(
    med_class_map,
    by = c("prev_medication" = "medication")
  ) %>%
  rename(prev_class = med_class) %>%
  left_join(
    med_class_map,
    by = c("medication" = "medication")
  ) %>%
  rename(curr_class = med_class) %>%
  filter(
    prev_class %in% c("Sedative", "Antipsychotic"),
    curr_class %in% c("Sedative", "Antipsychotic")
  )

pod_switches_class <- pod_switches_class %>%
  mutate(
    switch_class = case_when(
      prev_class == "Sedative" & curr_class == "Antipsychotic" ~
        "Sedative → Antipsychotic",
      prev_class == "Antipsychotic" & curr_class == "Sedative" ~
        "Antipsychotic → Sedative",
      prev_class == "Antipsychotic" & curr_class == "Antipsychotic" ~
        "Antipsychotic → Antipsychotic",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(switch_class))

switch_patient_level <- pod_switches_class %>%
  group_by(subject_id, cluster, switch_class) %>%
  summarise(any_switch = TRUE, .groups = "drop")

switch_heatmap_data <- switch_patient_level %>%
  group_by(cluster) %>%
  mutate(
    n_switchers = n_distinct(subject_id)
  ) %>%
  ungroup() %>%
  group_by(cluster, switch_class, n_switchers) %>%
  summarise(
    percent = 100 * n() / n_switchers,
    .groups = "drop"
  )

medication_switch_heatmap <- ggplot(
  switch_heatmap_data,
  aes(
    x = factor(cluster),
    y = switch_class,
    fill = percent
  )
) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(
  aes(label = sprintf("%.1f%%", percent)),
  size  = 3.5,
  color = "white",
  fontface = "bold"
)+
  scale_fill_viridis_c(
    option = "viridis",
    name = "% of switching patients",
    limits = c(0, 100)
  ) +
  coord_fixed() +
  labs(
    x = "Cluster",
    y = "Medication switch class"
  ) +
  theme_nature +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 0, vjust = 0.5)
  )+
  theme(
    panel.grid = element_blank(),

    # Shrink side labels
    axis.text.x  = element_text(size = 8),
    axis.text.y  = element_text(size = 8),
    axis.title.x = element_text(size = 8),
    axis.title.y = element_text(size = 8),

    # Legend smaller
    legend.title = element_text(size = 8),
    legend.text  = element_text(size = 7),

    # Pull plot outward
    plot.margin = margin(
      t = 6, r = 6, b = 6, l = 6, unit = "pt"
    )
  )

medication_switch_heatmap

ggsave(
  "figures/main/01_POD/pod_medication_switch_class_among_switching_patients_heatmap.png",
  medication_switch_heatmap,
  width  = 7,
  height = 3.6,
  dpi    = 300,
  bg     = "white",
  device = ragg::agg_png)
```

## Medication Type Switches (Table)

```{r, echo=FALSE, message=FALSE}
med_class_map <- tibble(
  medication = c(
    "Dexmedetomidine",
    "Quetiapine",
    "Haloperidol",
    "Olanzapine",
    "Risperidone",
    "Chlorpromazine",
    "Rivastigmine"
  ),
  med_class = c(
    "Sedative",          # Dexmedetomidine
    "Antipsychotic",
    "Antipsychotic",
    "Antipsychotic",
    "Antipsychotic",
    "Antipsychotic",
    "Other"              # Rivastigmine (exclude later)
  )
)

pod_switches_class <- pod_switches %>%
  filter(is_switch) %>%
  left_join(
    med_class_map,
    by = c("prev_medication" = "medication")
  ) %>%
  rename(prev_class = med_class) %>%
  left_join(
    med_class_map,
    by = c("medication" = "medication")
  ) %>%
  rename(curr_class = med_class) %>%
  filter(
    prev_class %in% c("Sedative", "Antipsychotic"),
    curr_class %in% c("Sedative", "Antipsychotic")
  )

pod_switches_class <- pod_switches_class %>%
  mutate(
    switch_class = case_when(
      prev_class == "Sedative" & curr_class == "Antipsychotic" ~
        "Sedative → Antipsychotic",
      prev_class == "Antipsychotic" & curr_class == "Sedative" ~
        "Antipsychotic → Sedative",
      prev_class == "Antipsychotic" & curr_class == "Antipsychotic" ~
        "Antipsychotic → Antipsychotic",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(switch_class))

switch_patient_level <- pod_switches_class %>%
  group_by(subject_id, cluster, switch_class) %>%
  summarise(any_switch = TRUE, .groups = "drop")

switch_summary <- switch_patient_level %>%
  group_by(cluster) %>%
  mutate(n_switchers = n_distinct(subject_id)) %>%
  ungroup() %>%
  group_by(cluster, switch_class) %>%
  summarise(
    n = n_distinct(subject_id),
    n_switchers = first(n_switchers),
    .groups = "drop"
  ) %>%
  mutate(
    pct = 100 * n / n_switchers,
    value = sprintf("%d (%.1f%%)", n, pct)
  )

switch_table <- switch_summary %>%
  select(cluster, switch_class, value) %>%
  pivot_wider(
    names_from = switch_class,
    values_from = value,
    values_fill = "0 (0.0%)"
  ) %>%
  arrange(cluster)

# Rebame columns 
colnames(switch_table) <- c("Cluster", 
                            "Antipsychotic → Antipsychotic",
                            "Antipsychotic → Sedative",
                            "Sedative → Antipsychotic")

# Rearrange the column order
switch_table <- switch_table %>% select(Cluster, `Sedative → Antipsychotic`, `Antipsychotic → Sedative`, `Antipsychotic → Antipsychotic`)

# Ensure named vector for colors
cluster_color_map <- cluster_colors[2:5]
names(cluster_color_map) <- c(2,3,4,5)

cluster_switch_table_kable <- switch_table %>%
  kable(
    format = "html",
    escape = FALSE,
    align = "lccc"
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = NULL,
    font_size = 14
  ) %>%

  # Fix column widths (this enables wrapping)
  column_spec(1, width = "1.5cm") %>%   # Cluster
  column_spec(2, width = "3.5cm") %>% 
  column_spec(3, width = "3.5cm") %>%
  column_spec(4, width = "3.5cm") %>%  
  column_spec(
    1,
    bold = TRUE,
    background = cluster_color_map[cluster_summary$Cluster],
    color = "black"
  )

cluster_switch_table_kable

# Save table as a PNG
save_kable(
  cluster_switch_table_kable,
  file = "figures/main/01_POD/pod_medication_switch_class_among_switching_patients_table.png",
  zoom = 2
)
```

## Sedative vs. Antipsychotic Percentages by Cluster 

```{r, echo=FALSE, message=FALSE}

pod_medications_filtered <- pod_medications_filtered %>% mutate(drug_class = case_when(medication == "Dexmedetomidine" ~ "Sedative", medication %in% c("Quetiapine", "Haloperidol", "Olanzapine", "Risperidone", "Chlorpromazine") ~ "Antipsychotic", TRUE ~ "Other"))

gender_summary <- pod_study_cohort_features %>%
  count(cluster, gender, name = "n_patients") %>%
  group_by(cluster) %>%
  mutate(
    percent = 100 * n_patients / sum(n_patients)
  ) %>%
  ungroup()

# Collapse to patient level exposure 
patient_class_exposure <- pod_medications_filtered %>%
  filter(drug_class %in% c("Antipsychotic", "Sedative")) %>%
  distinct(subject_id, cluster, drug_class)

cluster_totals <- pod_study_cohort_features %>%
  distinct(subject_id, cluster) %>%
  count(cluster, name = "n_cluster")

# Compute percentages of patients in each cluster
class_summary <- patient_class_exposure %>%
  count(cluster, drug_class, name = "n_patients") %>%
  left_join(cluster_totals, by = "cluster") %>%
  mutate(
    percent = 100 * n_patients / n_cluster
  )

# Create bar plot for percentage of patients in each cluster who receive >= 1 antipsychotic and >= 1 sedative drug 
med_class_plot <- ggplot(
  class_summary,
  aes(
    x = factor(cluster),
    y = percent,
    fill = drug_class
  )
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.55
  ) +
  geom_text(
    aes(
      label = paste0(
        sprintf("%.1f%%", percent),
        "\n(n=", n_patients, ")"
      )
    ),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 2.65,
    lineheight = 0.9
  ) +
scale_fill_manual(
  values = c(
    "Antipsychotic" = "#1D3557", 
    "Sedative"      = "#F0E442"   
  )
)+
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    x = "Cluster",
    y = "Percentage of patients",
    fill = "Medication class"
  ) +
  theme_nature

med_class_plot

ggsave(
  "figures/main/01_POD/pod_sedative_vs_antipsychotic_bar_plot.png",
  med_class_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

## Drug Administrations Before/After Delirium Onset

```{r, echo=FALSE, message=FALSE}

# Classify medications as before vs. after onset
pod_timing <- pod_medications_filtered %>%
  mutate(
    timing = factor(
      case_when(
        starttime < first_cam_positive_time  ~ "Before delirium onset",
        starttime >= first_cam_positive_time ~ "After delirium onset"
      ),
      levels = c(
        "Before delirium onset",
        "After delirium onset")))

# Compute percentages within each cluster
timing_summary <- pod_timing %>%
  count(cluster, timing, name = "n_admin") %>%
  group_by(cluster) %>%
  mutate(
    percent = 100 * n_admin / sum(n_admin)
  ) %>%
  ungroup()

# Create bar plot
admin_timing_plot <- ggplot(
  timing_summary,
  aes(
    x = factor(cluster),
    y = percent,
    fill = timing
  )
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.6
  ) +
  geom_text(
    aes(label = sprintf("%.1f%%", percent)),
    position = position_dodge(width = 0.9),
    vjust = -0.4,
    size = 2.8
  ) +
  scale_fill_manual(
  values = c(
    "Before delirium onset" = "#1D3557", 
    "After delirium onset"      = "#F0E442"   
  )
)+
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.08))
  ) +
  labs(
    x = "Cluster",
    y = "Percentage of medication administrations",
    fill = ""
  ) +
  theme_nature

admin_timing_plot

ggsave(
  "figures/main/01_POD/pod_medication_admin_timing_bar_plot.png",
  admin_timing_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## Dexmedetomidine Dose Analysis (Entire ICU Stay)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Note: Dexmedetomidine has units of mg and mcg 

# Dexmedetomidine exposure bar plot (compute the cumulative dose per patients exposed and normalize by length of ICU stay)
dexmed <- pod_medications_filtered %>%
  filter(medication == "Dexmedetomidine") %>%
  mutate(
    amount_mcg = case_when(
      units == "mcg" ~ amount,
      units == "mg"  ~ amount * 1000,
      TRUE ~ NA_real_
    )
  ) %>%
  filter(!is.na(amount_mcg))

dexmed_patient <- dexmed %>%
  group_by(subject_id, cluster, los_days) %>%
  summarise(
    cumulative_dose_mcg = sum(amount_mcg),
    dose_per_day_mcg    = cumulative_dose_mcg / los_days,
    .groups = "drop"
  )

dexmed_cluster_summary <- dexmed_patient %>%
  group_by(cluster) %>%
  summarise(
    mean_dose_per_day = mean(dose_per_day_mcg),
    sd_dose_per_day   = sd(dose_per_day_mcg),
    n_patients        = n(),
    .groups = "drop"
  )

# Bar plot
ggplot(
  dexmed_cluster_summary,
  aes(
    x = factor(cluster),
    y = mean_dose_per_day,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_dose_per_day - sd_dose_per_day / sqrt(n_patients),
      ymax = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients)
    ),
    width = 0.2
  ) +
  geom_text(
  aes(
    y = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients),
    label = sprintf("%.1f", mean_dose_per_day)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Mean dexmedetomidine dose (mcg/day)",
    title = "Dexmedetomidine mean daily doses",
    subtitle = "Only treated patients included"
  ) +
  theme_nature
```

## Quetiapine Dose Analysis (Entire ICU Stay)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Note: Quetiapine only has units of mg 

# Quetiapine dose bar plot (compute the cumulative dose per patients exposed and normalize by length of ICU stay)
quetiapine <- pod_medications_filtered %>%
  filter(medication == "Quetiapine")

quetiapine_patient <- quetiapine %>%
  group_by(subject_id, cluster, los_days) %>%
  summarise(
    cumulative_dose_mg = sum(amount),
    dose_per_day_mg    = cumulative_dose_mg / los_days,
    .groups = "drop"
  )

quetiapine_cluster_summary <- quetiapine_patient %>%
  group_by(cluster) %>%
  summarise(
    mean_dose_per_day = mean(dose_per_day_mg),
    sd_dose_per_day   = sd(dose_per_day_mg),
    n_patients        = n(),
    .groups = "drop"
  )

# Bar plot
ggplot(
  quetiapine_cluster_summary,
  aes(
    x = factor(cluster),
    y = mean_dose_per_day,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_dose_per_day - sd_dose_per_day / sqrt(n_patients),
      ymax = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients),
    label = sprintf("%.1f", mean_dose_per_day)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Mean quetiapine dose (mg/day)",
    title = "Quetiapine mean daily doses",
    subtitle = "Only treated patients included"
  ) +
  theme_nature
```

## Haloperidol Dose Analysis (Entire ICU Stay)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Note: Haloperidol only has units of mg 

# Haloperidol dose bar plot (compute the cumulative dose per patients exposed and normalize by length of ICU stay)
haloperidol <- pod_medications_filtered %>%
  filter(medication == "Haloperidol")

haloperidol_patient <- haloperidol %>%
  group_by(subject_id, cluster, los_days) %>%
  summarise(
    cumulative_dose_mg = sum(amount),
    dose_per_day_mg    = cumulative_dose_mg / los_days,
    .groups = "drop"
  )

haloperidol_cluster_summary <- haloperidol_patient %>%
  group_by(cluster) %>%
  summarise(
    mean_dose_per_day = mean(dose_per_day_mg),
    sd_dose_per_day   = sd(dose_per_day_mg),
    n_patients        = n(),
    .groups = "drop"
  )

# Bar plot
ggplot(
  haloperidol_cluster_summary,
  aes(
    x = factor(cluster),
    y = mean_dose_per_day,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_dose_per_day - sd_dose_per_day / sqrt(n_patients),
      ymax = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients),
    label = sprintf("%.1f", mean_dose_per_day)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Mean haloperidol dose (mg/day)",
    title = "Haloperidol mean daily doses",
    subtitle = "Only treated patients included"
  ) +
  theme_nature
```

## Dexmedetomidine Dose Analysis (Between first positive CAM-ICU and delirium resolution)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Note: Dexmedetomidine has units of mg and mcg 

# Dexmedetomidine exposure bar plot (compute the cumulative dose per patients exposed and normalize by length of ICU stay)
dexmed <- pod_medications_filtered %>%
  filter(medication == "Dexmedetomidine") %>%
  filter(starttime >= first_cam_positive_time & starttime <= delirium_endtime) %>%
  mutate(
    amount_mcg = case_when(
      units == "mcg" ~ amount,
      units == "mg"  ~ amount * 1000,
      TRUE ~ NA_real_
    )
  ) %>%
  filter(!is.na(amount_mcg))

dexmed_patient <- dexmed %>%
  group_by(subject_id, cluster, pod_resolution_days) %>%
  summarise(
    cumulative_dose_mcg = sum(amount_mcg),
    dose_per_day_mcg    = cumulative_dose_mcg / pod_resolution_days,
    .groups = "drop"
  )

dexmed_cluster_summary <- dexmed_patient %>%
  group_by(cluster) %>%
  summarise(
    mean_dose_per_day = mean(dose_per_day_mcg),
    sd_dose_per_day   = sd(dose_per_day_mcg),
    n_patients        = n(),
    .groups = "drop"
  )

# Bar plot
dexmed_dose_plot <- ggplot(
  dexmed_cluster_summary,
  aes(
    x = factor(cluster),
    y = mean_dose_per_day,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_dose_per_day - sd_dose_per_day / sqrt(n_patients),
      ymax = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients),
    label = sprintf("%.1f", mean_dose_per_day)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Mean cumulative dexmedetomidine dose (mcg/day)",
    #title = "Dexmedetomidine mean daily doses during delirium",
    #subtitle = "Only treated patients included"
  ) +
  theme_nature

dexmed_dose_plot

ggsave(
  "figures/main/01_POD/pod_mean_cumulative_daily_dexmedetomidine_dose_during_delirium_bar_plot.png",
  dexmed_dose_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## Quetiapine Dose Analysis (Between first positive CAM-ICU and delirium resolution)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Note: Quetiapine only has units of mg 

# Quetiapine dose bar plot (compute the cumulative dose per patients exposed and normalize by length of ICU stay)
quetiapine <- pod_medications_filtered %>%
  filter(medication == "Quetiapine") %>%
  filter(starttime >= first_cam_positive_time & starttime <= delirium_endtime)

quetiapine_patient <- quetiapine %>%
  group_by(subject_id, cluster, pod_resolution_days) %>%
  summarise(
    cumulative_dose_mg = sum(amount),
    dose_per_day_mg    = cumulative_dose_mg / pod_resolution_days,
    .groups = "drop"
  )

quetiapine_cluster_summary <- quetiapine_patient %>%
  group_by(cluster) %>%
  summarise(
    mean_dose_per_day = mean(dose_per_day_mg),
    sd_dose_per_day   = sd(dose_per_day_mg),
    n_patients        = n(),
    .groups = "drop"
  )

# Bar plot
quetiapine_dose_plot <- ggplot(
  quetiapine_cluster_summary,
  aes(
    x = factor(cluster),
    y = mean_dose_per_day,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_dose_per_day - sd_dose_per_day / sqrt(n_patients),
      ymax = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients),
    label = sprintf("%.1f", mean_dose_per_day)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Mean cumulative quetiapine dose (mg/day)",
    #title = "Quetiapine mean daily doses during delirium",
    #subtitle = "Only treated patients included"
  ) +
  theme_nature

quetiapine_dose_plot

ggsave(
  "figures/main/01_POD/pod_mean_cumulative_daily_quetiapine_dose_during_delirium_bar_plot.png",
  quetiapine_dose_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## Haloperidol Dose Analysis (Between first positive CAM-ICU and delirium resolution)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Note: Haloperidol only has units of mg 

# Haloperidol dose bar plot (compute the cumulative dose per patients exposed and normalize by length of ICU stay)
haloperidol <- pod_medications_filtered %>%
  filter(medication == "Haloperidol") %>%
  filter(starttime >= first_cam_positive_time & starttime <= delirium_endtime)

haloperidol_patient <- haloperidol %>%
  group_by(subject_id, cluster, pod_resolution_days) %>%
  summarise(
    cumulative_dose_mg = sum(amount),
    dose_per_day_mg    = cumulative_dose_mg / pod_resolution_days,
    .groups = "drop"
  )

haloperidol_cluster_summary <- haloperidol_patient %>%
  group_by(cluster) %>%
  summarise(
    mean_dose_per_day = mean(dose_per_day_mg),
    sd_dose_per_day   = sd(dose_per_day_mg),
    n_patients        = n(),
    .groups = "drop"
  )

# Bar plot
haloperidol_dose_plot <- ggplot(
  haloperidol_cluster_summary,
  aes(
    x = factor(cluster),
    y = mean_dose_per_day,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_dose_per_day - sd_dose_per_day / sqrt(n_patients),
      ymax = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_dose_per_day + sd_dose_per_day / sqrt(n_patients),
    label = sprintf("%.1f", mean_dose_per_day)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Mean cumulative haloperidol dose (mg/day)",
    #title = "Haloperidol mean daily doses during delirium",
    #subtitle = "Only treated patients included"
  ) +
  theme_nature

haloperidol_dose_plot

ggsave(
  "figures/main/01_POD/pod_mean_cumulative_daily_haloperidol_dose_during_delirium_bar_plot.png",
  haloperidol_dose_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

# Demographics and Clinical Risk Factors

## Age

```{r, echo=FALSE, message=FALSE}
age_summary <- pod_study_cohort_features %>%
  group_by(cluster) %>%
  summarise(
    mean_age = mean(age),
    sd_age   = sd(age),
    n_patients      = n(),
    .groups = "drop"
  )

ggplot(
  age_summary,
  aes(
    x = factor(cluster),
    y = mean_age,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_age - sd_age / sqrt(n_patients),
      ymax = mean_age + sd_age / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_age + sd_age / sqrt(n_patients),
    label = sprintf("%.1f", mean_age)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    x = "Cluster",
    y = "Mean age (years)",
  ) +
  theme_nature
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(
  pod_study_cohort_features,
  aes(x = age, fill = factor(cluster))
) +
  geom_histogram(bins = 12, alpha = 0.8, color = "white") +
  facet_wrap(~ cluster, scales = "free_y") +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Age (years)",
    y = "Number of patients"
  ) +
  theme_nature
```

## Gender

```{r, echo=FALSE, message=FALSE}
gender_summary <- pod_study_cohort_features %>%
  count(cluster, gender, name = "n_patients") %>%
  group_by(cluster) %>%
  mutate(
    percent = 100 * n_patients / sum(n_patients)
  ) %>%
  ungroup()

gender_plot <- ggplot(
  gender_summary,
  aes(
    x = factor(cluster),
    y = percent,
    fill = gender
  )
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.55
  ) +
geom_text(
  aes(
    label = paste0(
      sprintf("%.1f%%", percent),
      "\n(n=", n_patients, ")"
    )
  ),
  position = position_dodge(width = 0.9),
  vjust = -0.4,
  size = 2.65,
  lineheight = 0.9
) +
    scale_fill_manual(
  values = c(
    "Female" = "#1D3557", 
    "Male"      = "#F0E442"   
  )
)+
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    x = "Cluster",
    y = "Percentage of patients",
    fill = "Gender"
  ) +
  theme_nature

gender_plot

ggsave(
  "figures/main/01_POD/pod_gender_bar_plot.png",
  gender_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

## Alcohol use disorder 

```{r, echo=FALSE, message=FALSE}
alcohol_use_disorder_summary <- pod_study_cohort_features %>%
  count(cluster, alcohol_use_disorder, name = "n_patients") %>%
  group_by(cluster) %>%
  mutate(
    percent = 100 * n_patients / sum(n_patients)
  ) %>%
  ungroup()

alcohol_use_disorder_plot <- ggplot(
  alcohol_use_disorder_summary,
  aes(
    x = factor(cluster),
    y = percent,
    fill = alcohol_use_disorder
  )
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.6
  ) +
geom_text(
  aes(
    label = paste0(
      sprintf("%.1f%%", percent),
      "\n(n=", n_patients, ")"
    )
  ),
  position = position_dodge(width = 0.9),
  vjust = -0.4,
  size = 2.6,
  lineheight = 0.9
) +
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.15))
  ) +
    scale_fill_manual(
  values = c(
    "No" = "#1D3557", 
    "Yes"      = "#F0E442"   
  )
)+
  labs(
    x = "Cluster",
    y = "Percentage of patients",
    fill = "Alcohol use disorder"
  ) +
  theme_nature

alcohol_use_disorder_plot

ggsave(
  "figures/main/01_POD/pod_alcohol_use_disorder_bar_plot.png",
  alcohol_use_disorder_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

## Mechanical Ventilator Use

```{r, echo=FALSE, message=FALSE}
ventilator_use_summary <- pod_study_cohort_features %>%
  count(cluster, ventilator_use, name = "n_patients") %>%
  group_by(cluster) %>%
  mutate(
    percent = 100 * n_patients / sum(n_patients)
  ) %>%
  ungroup()

ventilator_use_plot <- ggplot(
  ventilator_use_summary,
  aes(
    x = factor(cluster),
    y = percent,
    fill = ventilator_use
  )
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.55
  ) +
geom_text(
  aes(
    label = paste0(
      sprintf("%.1f%%", percent),
      "\n(n=", n_patients, ")"
    )
  ),
  position = position_dodge(width = 0.9),
  vjust = -0.4,
  size = 2.6,
  lineheight = 0.9
) +
    scale_fill_manual(
  values = c(
    "No" = "#1D3557", 
    "Yes"      = "#F0E442"   
  )
)+
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    x = "Cluster",
    y = "Percentage of patients",
    fill = "Ventilator use"
  ) +
  theme_nature

ventilator_use_plot

ggsave(
  "figures/main/01_POD/pod_ventilator_use_bar_plot.png",
  ventilator_use_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

## Mechanical Ventilation Duration (among ventilated patients)

```{r, echo=FALSE, message=FALSE}
ventilated_patients <- pod_study_cohort_features %>%
  filter(ventilation_duration_mins > 0)

vent_summary <- ventilated_patients %>%
  group_by(cluster) %>%
  summarise(
    mean_vent_days = mean(ventilation_duration_mins / (60*24), na.rm = TRUE),
    sd_vent_days   = sd(ventilation_duration_mins / (60*24), na.rm = TRUE),
    n_patients      = n(),
    .groups = "drop"
  )

ventilation_duration_plot <- ggplot(
  vent_summary,
  aes(
    x = factor(cluster),
    y = mean_vent_days,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_vent_days - sd_vent_days / sqrt(n_patients),
      ymax = mean_vent_days + sd_vent_days / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_vent_days + sd_vent_days / sqrt(n_patients),
    label = sprintf("%.1f", mean_vent_days)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "Cluster",
    y = "Mean ventilation duration (days)",
    #title = "Mechanical ventilation duration",
    #subtitle = "Only ventilated patients included"
  ) +
  theme_nature

ventilation_duration_plot

ggsave(
  "figures/main/01_POD/pod_ventilation_duration_bar_plot.png",
  ventilation_duration_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}

ggplot(
  ventilated_patients,
  aes(x = ventilation_duration_mins / (60*24), fill = factor(cluster))
) +
  geom_histogram(bins = 30, alpha = 0.8, color = "white") +
  facet_wrap(~ cluster, scales = "free_y") +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_x_continuous(limits = c(0, 30), labels = c(0, 10, 20, 30)) +
  labs(
    x = "Ventilation duration (days)",
    y = "Number of patients"
  ) +
  theme_nature
```



# Health Outcomes 

## Discharge Location 

```{r, echo=FALSE, message=FALSE}
# Group discharge locations into common categories 
pod_discharge_location <- pod_study_cohort_features %>%
  mutate(
    discharge_location = case_when(
      discharge_location %in% c("HOME", "HOME HEALTH CARE") ~ "Home/HHC",
      discharge_location == "REHAB" ~ "Rehab",
      discharge_location == "SKILLED NURSING FACILITY" ~ "SNF",
      discharge_location == "CHRONIC/LONG TERM ACUTE CARE" ~ "LTACH",
      discharge_location %in% c(
        "PSYCH FACILITY", "OTHER FACILITY", "ASSISTED LIVING",
        "AGAINST ADVICE", "ACUTE HOSPITAL"
      ) ~ "Other",
      discharge_location == "HOSPICE" ~ "Hospice",
      discharge_location == "DIED" ~ "Died"
    ),
    discharge_location = factor(
      discharge_location,
      levels = c(
        "Died",
        "Hospice",
        "Other",
        "LTACH",
        "SNF",
        "Rehab",
        "Home/HHC"
      ),
      ordered = TRUE
    )
  )

discharge_pct <- pod_discharge_location %>%
  mutate(
    cluster = factor(cluster),
    discharge_location = factor(discharge_location)
  ) %>%
  group_by(cluster, discharge_location) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(
    pct = 100 * n / sum(n)
  )

discharge_locations_plot <- ggplot(
  discharge_pct,
  aes(x = discharge_location, y = pct, fill = cluster)
) +
  geom_col(width = 0.8) +
  facet_wrap(
  ~ cluster,
  labeller = labeller(cluster = function(x) paste("Cluster", x))
) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(
    #labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    x = "Discharge location",
    y = "Percentage of patients"
  ) +
  geom_text(
  aes(
    label = paste0(
      sprintf("%.1f%%", pct)),
    y = pct
  ),
  hjust = -0.05,
  size = 2.4
) +
  coord_flip() + theme_nature

discharge_locations_plot 

ggsave(
  "figures/main/01_POD/pod_discharge_locations_bar_plot.png",
  discharge_locations_plot,
  width  = 9,
  height = 4,
  dpi    = 300,
  bg     = "white")
```

## RASS Scores (At time of first positive CAM-ICU assessment)

```{r, echo=FALSE, message=FALSE}
# RASS Scores 
rass_pct <- rass_scores_filtered %>%
  filter(
    label == "Richmond-RAS Scale",
    charttime == first_cam_positive_time
  ) %>%
  mutate(
    cluster  = factor(cluster),
    valuenum = factor(valuenum, levels = -5:4)
  ) %>%
  group_by(cluster, valuenum) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(
    pct = 100 * n / sum(n)
  )

rass_scores_plot <- ggplot(
  rass_pct,
  aes(x = valuenum, y = pct, fill = cluster)
) +
  geom_col(width = 0.8) +
  facet_wrap(
  ~ cluster,
  labeller = labeller(cluster = function(x) paste("Cluster", x))
) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(
    #labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  labs(
    x = "RASS score",
    y = "Percentage of patients"
  ) +
  geom_text(
  aes(
    label = paste0(
      sprintf("%.0f%%", pct)),
    y = pct
  ),
  vjust = -0.25,
  size = 2.3
) +
  theme_nature

rass_scores_plot

ggsave(
  "figures/main/01_POD/pod_rass_scores_first_cam_icu_bar_plot.png",
  rass_scores_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## RASS Scores (After delirium onset)

```{r, echo=FALSE, message=FALSE}
# RASS Scores after delirium onset 
rass_pct <- rass_scores_filtered %>%
  filter(
    label == "Richmond-RAS Scale",
    charttime >= first_cam_positive_time
  ) %>%
  mutate(
    cluster  = factor(cluster),
    valuenum = factor(valuenum, levels = -5:4)
  ) %>%
  group_by(cluster, valuenum) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(
    pct = 100 * n / sum(n)
  )

rass_scores_plot <- ggplot(
  rass_pct,
  aes(x = valuenum, y = pct, fill = cluster)
) +
  geom_col(width = 0.8) +
  facet_wrap(
  ~ cluster,
  labeller = labeller(cluster = function(x) paste("Cluster", x))
) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(
    #labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    x = "RASS score",
    y = "Percentage of scores"
  ) +
  geom_text(
  aes(
    label = paste0(
      sprintf("%.0f%%", pct)),
    y = pct
  ),
  vjust = -0.25,
  size = 2.3
) +
  theme_nature

rass_scores_plot

ggsave(
  "figures/main/01_POD/pod_rass_scores_after_delirium_onset_bar_plot.png",
  rass_scores_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## RASS Scores (Between first positive CAM-ICU result and delirium resolution)

```{r, echo=FALSE, message=FALSE}
# RASS Scores after delirium onset 
rass_pct <- rass_scores_filtered %>%
  filter(
    label == "Richmond-RAS Scale",
    charttime >= first_cam_positive_time & charttime <= delirium_endtime
  ) %>%
  mutate(
    cluster  = factor(cluster),
    valuenum = factor(valuenum, levels = -5:4)
  ) %>%
  group_by(cluster, valuenum) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(
    pct = 100 * n / sum(n)
  )

rass_scores_plot <- ggplot(
  rass_pct,
  aes(x = valuenum, y = pct, fill = cluster)
) +
  geom_col(width = 0.8) +
  facet_wrap(
  ~ cluster,
  labeller = labeller(cluster = function(x) paste("Cluster", x))
) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(
    #labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    x = "RASS score",
    y = "Percentage of scores"
  ) +
  geom_text(
  aes(
    label = paste0(
      sprintf("%.0f%%", pct)),
    y = pct
  ),
  vjust = -0.25,
  size = 2.3
) +
  theme_nature

rass_scores_plot

ggsave(
  "figures/main/01_POD/pod_rass_scores_during_delirium_bar_plot.png",
  rass_scores_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## POD Duration 

```{r, echo=FALSE, message=FALSE}
pod_duration_summary <- pod_study_cohort_features %>%
  group_by(cluster) %>%
  summarise(
    mean_pod_duration_days = mean(pod_duration_days),
    sd_pod_duration_days   = sd(pod_duration_days),
    n_patients      = n(),
    .groups = "drop"
  )

ggplot(
  pod_duration_summary,
  aes(
    x = factor(cluster),
    y = mean_pod_duration_days,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_pod_duration_days - sd_pod_duration_days / sqrt(n_patients),
      ymax = mean_pod_duration_days + sd_pod_duration_days / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_pod_duration_days + sd_pod_duration_days / sqrt(n_patients),
    label = sprintf("%.1f", mean_pod_duration_days)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.17))) +
  labs(
    x = "Cluster",
    y = "Mean POD duration (days)",
  ) + theme_nature
```

## POD Resolution Time Summary Statistics

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# KM-estimated median POD resolution times by cluster
pod_study_cohort_features <- pod_study_cohort_features %>%
  mutate(event = !censored)

fit <- survfit(
  Surv(pod_resolution_days, event) ~ cluster,
  data = pod_study_cohort_features
)

kable(summary(fit)$table)

# Examine death as a competing risk
all_pod_study_cohort_features <- all_pod_study_cohort_features %>%
  mutate(
    death_before_resolution = case_when(
      censored == TRUE & !is.na(deathtime) & deathtime <= icu_outtime ~ 1,
      TRUE ~ 0
    )
  )

died_before_resolution <- all_pod_study_cohort_features %>% filter(death_before_resolution == TRUE)

sprintf("%s of patients died before POD resolution.", round(100*(nrow(died_before_resolution)/nrow(all_pod_study_cohort_features)),3))

# Cluster-level summary of death as a competing risk
pod_study_cohort_features <- pod_study_cohort_features %>%
  mutate(
    death_before_resolution = case_when(
      censored == TRUE & !is.na(deathtime) & deathtime <= icu_outtime ~ 1,
      TRUE ~ 0
    )
  )

death_summary <- pod_study_cohort_features %>%
  group_by(cluster) %>%
  summarise(
    n_total = n(),
    n_death_before_resolution = sum(death_before_resolution),
    percent_death_before_resolution = 100 * n_death_before_resolution / n_total,
    .groups = "drop"
  )

death_vs_censored <- pod_study_cohort_features %>%
  group_by(cluster) %>%
  summarise(
    n_censored = sum(censored),
    n_death_before_resolution = sum(death_before_resolution),
    percent_of_censored_due_to_death = 100 * n_death_before_resolution / n_censored,
    .groups = "drop"
  )

kable(death_summary)
kable(death_vs_censored)
```

## POD Resolution Time Kaplan-Meier Curve

```{r, echo=FALSE, message=FALSE}
# KM-curve with global log-rank test (with 95% confidence bands)

updated_cluster_colors <- cluster_colors
names(updated_cluster_colors) <- paste("Cluster", names(cluster_colors))

surv_data <- pod_study_cohort_features %>%
  mutate(
    time    = pod_resolution_days,
    event   = as.integer(!censored),  # 1 = resolved, 0 = censored
    cluster = factor(
      cluster,
      labels = paste("Cluster", sort(unique(cluster)))
    )
  )


km_fit <- survfit(
  Surv(time, event) ~ cluster,
  data = surv_data
)

km_df <- data.frame(
  time    = km_fit$time,
  surv    = km_fit$surv,
  cluster = rep(names(km_fit$strata), km_fit$strata)
) %>%
  mutate(
    cluster = str_remove(cluster, "^cluster=")
  )

logrank_test <- survdiff(
  Surv(time, event) ~ cluster,
  data = surv_data
)

# Extract p-value
logrank_p <- 1 - pchisq(
  logrank_test$chisq,
  df = length(logrank_test$n) - 1
)

logrank_label <- ifelse(
  logrank_p < 0.001,
  "Log-rank p < 0.001",
  paste0("Log-rank p = ", formatC(logrank_p, digits = 3, format = "f"))
)

km_curve <- ggplot(km_df, aes(x = time, y = surv, color = cluster)) +
  geom_step(linewidth = 0.8) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  scale_color_manual(values = updated_cluster_colors) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = logrank_label,
    hjust = 1.1,
    vjust = 1.5,
    size = 3.6
  ) +
  labs(
    x = "Days since first positive CAM-ICU",
    y = "Probability of unresolved delirium",
    color = "Cluster"
  ) +
  theme_nature

km_curve

ggsave(
  "figures/main/01_POD/pod_resolution_time_kaplan_meier_curve.png",
  km_curve,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## POD Resolution Time Kaplan-Meier Curve (with confidence intervals)

```{r, echo=FALSE, message=FALSE}
# KM-curve with global log-rank test 

updated_cluster_colors <- cluster_colors
names(updated_cluster_colors) <- paste("Cluster", names(cluster_colors))

surv_data <- pod_study_cohort_features %>%
  mutate(
    time    = pod_resolution_days,
    event   = as.integer(!censored),  # 1 = resolved, 0 = censored
    cluster = factor(
      cluster,
      labels = paste("Cluster", sort(unique(cluster)))
    )
  )


km_fit <- survfit(
  Surv(time, event) ~ cluster,
  data = surv_data
)

km_df <- data.frame(
  time    = km_fit$time,
  surv    = km_fit$surv,
  lower   = km_fit$lower,
  upper   = km_fit$upper,
  cluster = rep(names(km_fit$strata), km_fit$strata)
) %>%
  mutate(
    cluster = str_remove(cluster, "^cluster=")
  )

logrank_test <- survdiff(
  Surv(time, event) ~ cluster,
  data = surv_data
)

# Extract p-value
logrank_p <- 1 - pchisq(
  logrank_test$chisq,
  df = length(logrank_test$n) - 1
)

logrank_label <- ifelse(
  logrank_p < 0.001,
  "Log-rank p < 0.001",
  paste0("Log-rank p = ", formatC(logrank_p, digits = 3, format = "f"))
)

km_curve <- ggplot(km_df, aes(x = time, y = surv, color = cluster)) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper, fill = cluster),
    alpha = 0.15,
    color = NA
  ) +
  geom_step(linewidth = 0.8) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  scale_color_manual(values = updated_cluster_colors) +
  scale_fill_manual(values = updated_cluster_colors) +
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = logrank_label,
    hjust = 1.1,
    vjust = 1.5,
    size = 3.6
  ) +
  labs(
    x = "Days since first positive CAM-ICU",
    y = "Probability of unresolved delirium",
    color = "Cluster",
    fill = "Cluster"
  ) +
  theme_nature

km_curve

ggsave(
  "figures/main/01_POD/pod_resolution_time_kaplan_meier_curve_with_confidence_intervals.png",
  km_curve,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## Recurrent POD

```{r, echo=FALSE, message=FALSE}
recurrent_summary <- pod_study_cohort_features %>%
  filter(recurrent %in% c("Yes", "No")) %>%   # safety check
  count(cluster, recurrent, name = "n_patients") %>%
  group_by(cluster) %>%
  mutate(
    percent = 100 * n_patients / sum(n_patients)
  ) %>%
  ungroup()

recurrent_pod_plot <- ggplot(
  recurrent_summary,
  aes(
    x = factor(cluster),
    y = percent,
    fill = recurrent
  )
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.5
  ) +
geom_text(
  aes(
    label = paste0(
      sprintf("%.1f%%", percent),
      "\n(n=", n_patients, ")"
    )
  ),
  position = position_dodge(width = 0.9),
  vjust = -0.4,
  size = 2.4,
  lineheight = 0.9
) +
    scale_fill_manual(
  values = c(
    "No" = "#1D3557", 
    "Yes"      = "#F0E442"   
  )
)+
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    x = "Cluster",
    y = "Percentage of patients",
    fill = "Recurrent delirium"
  ) + theme_nature

recurrent_pod_plot

ggsave(
  "figures/main/01_POD/pod_recurrent_vs_nonrecurrent_bar_plot.png",
  recurrent_pod_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

## In-Hospital Mortality 

```{r, echo=FALSE, message=FALSE}
mortality_summary <- pod_study_cohort_features %>%
  count(cluster, mortality, name = "n_patients") %>%
  group_by(cluster) %>%
  mutate(
    percent = 100 * n_patients / sum(n_patients)
  ) %>%
  ungroup()

in_hospital_mortality_plot <- ggplot(
  mortality_summary,
  aes(
    x = factor(cluster),
    y = percent,
    fill = mortality
  )
) +
  geom_col(
    position = position_dodge(width = 0.9),
    width = 0.5
  ) +
geom_text(
  aes(
    label = paste0(
      sprintf("%.1f%%", percent),
      "\n(n=", n_patients, ")"
    )
  ),
  position = position_dodge(width = 0.9),
  vjust = -0.4,
  size = 2.5,
  lineheight = 0.9
) +
    scale_fill_manual(
  values = c(
    "No" = "#1D3557", 
    "Yes"      = "#F0E442"   
  )
)+
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.01))
  ) +
  labs(
    x = "Cluster",
    y = "Percentage of patients",
    fill = "In-hospital mortality"
  ) + theme_nature 

in_hospital_mortality_plot

ggsave(
  "figures/main/01_POD/pod_in_hospital_mortality_bar_plot.png",
  in_hospital_mortality_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

## ICU Length of Stay 

```{r, echo=FALSE, message=FALSE}
los_summary <- pod_study_cohort_features %>%
  group_by(cluster) %>%
  summarise(
    mean_los_days = mean(los_days),
    sd_los_days   = sd(los_days),
    n_patients      = n(),
    .groups = "drop"
  )

icu_length_of_stay_plot <- ggplot(
  los_summary,
  aes(
    x = factor(cluster),
    y = mean_los_days,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_los_days - sd_los_days / sqrt(n_patients),
      ymax = mean_los_days + sd_los_days / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_los_days + sd_los_days / sqrt(n_patients),
    label = sprintf("%.1f", mean_los_days)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "Cluster",
    y = "Mean ICU length of stay (days)",
  ) + theme_nature 

icu_length_of_stay_plot

ggsave(
  "figures/main/01_POD/pod_icu_length_of_stay_bar_plot.png",
  icu_length_of_stay_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

# Hypothesis Testing 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Perform hypothesis tests to compare subgroups with Benjamini-Hochberg FDR adjusted p-values

# Ensure that cluster is a factor variable
pod_study_cohort_features$cluster <- factor(pod_study_cohort_features$cluster)

# Mechanical ventilation proportions (Chi-square)
mv_chisq <- chisq.test(table(pod_study_cohort_features$ventilator_use, pod_study_cohort_features$cluster))

p_mv_chisq <- mv_chisq$p.value

# Mechanical ventilation duration among ventilated patients (Kruskal-Wallis)
ventilated_patients <- pod_study_cohort_features %>%
  filter(ventilation_duration_mins > 0) %>%
  mutate(ventilation_duration_days = ventilation_duration_mins /(60*24))

mv_duration_kw <- kruskal.test(ventilation_duration_days ~ cluster, data = ventilated_patients)

p_mv_duration_kw <- mv_duration_kw$p.value

# Recurrent delirium (Chi-square)
recurrent_delirium_chisq <- chisq.test(table(pod_study_cohort_features$recurrent, pod_study_cohort_features$cluster))

p_recurrent_delirium_chisq <- recurrent_delirium_chisq$p.value

# RASS at delirium onset (Kruskal-Wallis)
rass_onset_scores <- rass_scores_filtered %>%
  filter(
    label == "Richmond-RAS Scale",
    charttime == first_cam_positive_time
  ) %>%
  mutate(
    cluster  = factor(cluster),
    valuenum = as.numeric(valuenum))

rass_kw <- kruskal.test(valuenum ~ cluster, data = rass_onset_scores)

p_rass_kw <- rass_kw$p.value

# Discharge location (Chi-square)
pod_discharge_location <- pod_study_cohort_features %>%
  mutate(
    discharge_location = case_when(
      discharge_location %in% c("HOME", "HOME HEALTH CARE") ~ "Home/HHC",
      discharge_location == "REHAB" ~ "Rehab",
      discharge_location == "SKILLED NURSING FACILITY" ~ "SNF",
      discharge_location == "CHRONIC/LONG TERM ACUTE CARE" ~ "LTACH",
      discharge_location %in% c(
        "PSYCH FACILITY", "OTHER FACILITY", "ASSISTED LIVING",
        "AGAINST ADVICE", "ACUTE HOSPITAL"
      ) ~ "Other",
      discharge_location == "HOSPICE" ~ "Hospice",
      discharge_location == "DIED" ~ "Died"
    ),
    discharge_location = factor(
      discharge_location,
      levels = c(
        "Died",
        "Hospice",
        "Other",
        "LTACH",
        "SNF",
        "Rehab",
        "Home/HHC"
      ),
      ordered = FALSE
    )
  )

discharge_chisq <- chisq.test(table(pod_discharge_location$discharge_location, pod_discharge_location$cluster))

p_discharge_chisq <- discharge_chisq$p.value

# Delirium resolution (global log-rank)
surv_data <- pod_study_cohort_features %>%
  mutate(
    time    = pod_resolution_days,
    event   = as.integer(!censored),  # 1 = resolved, 0 = censored
    cluster = factor(
      cluster,
      labels = paste("Cluster", sort(unique(cluster)))
    )
  )

km_fit <- survfit(
  Surv(time, event) ~ cluster,
  data = surv_data
)

km_df <- data.frame(
  time    = km_fit$time,
  surv    = km_fit$surv,
  cluster = rep(names(km_fit$strata), km_fit$strata)
) %>%
  mutate(
    cluster = str_remove(cluster, "^cluster=")
  )

logrank_test <- survdiff(
  Surv(time, event) ~ cluster,
  data = surv_data
)

logrank_p <- 1 - pchisq(
  logrank_test$chisq,
  df = length(logrank_test$n) - 1
)

# Calculate the BH adjusted p-values and display in table
# Format p-values
format_pval <- function(p) {
  ifelse(
    p < 0.001,
    "<0.001",
    sprintf("%.3f", p)
  )
}

# Calculate BH-adjusted p-values
test <- c(
  "Mechanical ventilation proportion (Chi-square)",
  "Mechanical ventilation duration among ventilated patients (Kruskal–Wallis)",
  "Recurrent delirium (Chi-square)",
  "RASS score at delirium onset (Kruskal–Wallis)",
  "Discharge location (Chi-square)",
  "Delirium resolution time (Global log-rank)"
)

raw_pval <- c(
  p_mv_chisq,
  p_mv_duration_kw,
  p_recurrent_delirium_chisq,
  p_rass_kw,
  p_discharge_chisq,
  logrank_p
)

p_adj <- p.adjust(raw_pval, method = "BH")

hypothesis_test_results <- data.frame(
  `Hypothesis test` = test,
  `Raw p-value` = format_pval(raw_pval),
  `BH-adjusted p-value` = format_pval(p_adj),
  check.names = FALSE
)

hypothesis_test_results <- hypothesis_test_results %>%
  mutate(`Hypothesis test` = str_replace_all(`Hypothesis test`, "Kruskal.{1,5}Wallis", "Kruskal-Wallis"))

kable(
  caption = "POD Application Hypothesis Testing Results",
  hypothesis_test_results,
  align = c("l", "c", "c")
)

# Save latex table
hypothesis_test_results_latex <- kable(
  hypothesis_test_results,
  format = "latex",
  booktabs = TRUE,
  longtable = FALSE,
  caption = "Hypothesis testing results for the POD application.",
  label = "tab:pod_hypothesis_tests",
  align = c("l", "c", "c"),
  escape = TRUE,
  linesep = ""
) %>%
  kable_styling(
    latex_options = c("hold_position"),
    position = "center"
  ) %>%
  column_spec(1, width = "9cm") %>%
  column_spec(2, width = "2.5cm") %>%
  column_spec(3, width = "3cm")

# Save as a .tex file
dir.create("tables/01_POD", recursive = TRUE, showWarnings = FALSE)
save_kable(hypothesis_test_results_latex,  "tables/01_POD/pod_hypothesis_test_results.tex")
```
