---
title: "PONV Application 4 windows GMM GT-base Text Template Cluster Analysis"
subtitle: "Cluster using the first 5 PCs from PCA, centered and unscaled, 8 clusters"
output: html_document
---

```{r setup, include=FALSE}
# Set working directory as project root directory 
knitr::opts_knit$set(root.dir = ".")
```

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Import libraries
library(readr)
library(tidyverse)
library(cluster)
library(clusterCrit)
library(uwot)
library(proxy)
library(kableExtra)
library(forcats)
library(lubridate)
library(arrow)
library(survival)

# Import data 
cluster_assignments <- read_csv("data/cohorts/02_PONV/GMM_clusters_4windows/ponv_gmm_4windows_gt_base_text_clusters_1rep_5pcs_centered_unscaled_8clusters.csv")
ponv_study_cohort_features <- read_csv("data/cohorts/02_PONV/ponv_study_cohort_features.csv")
ponv_medications <- read_csv("data/derived/02_PONV/ponv_medications_cleaned.csv")
embeddings <- read_parquet("data/embeddings/02_PONV/ponv_gatortron_base_text_template_concat4w_meanpool_l2norm.parquet")
```

## Data Cleaning 

```{r, echo=FALSE, message=FALSE}
# Specify the clustering method
method <- "gmm"

# Add cluster assignment column to ponv_study_cohort_features 
ponv_study_cohort_features <- left_join(ponv_study_cohort_features, cluster_assignments, by = "subject_id")

# Add cluster assignment column to ponv_medications
ponv_medications <- left_join(ponv_medications, cluster_assignments, by = "subject_id")

# Add admin_interval column to ponv_medications
ponv_medications <- ponv_medications %>% mutate(
  admin_interval = case_when(
    difftime(admin_time, procedure_date, units="days") >= -1 & difftime(admin_time, procedure_date, units="days") <= 0 ~ "-1 to 0 days",
    difftime(admin_time, procedure_date, units="days") > 0 & difftime(admin_time, procedure_date, units="days") <= 1 ~ "0 to 1 days",
    difftime(admin_time, procedure_date, units="days") > 1 & difftime(admin_time, procedure_date, units="days") <= 2 ~ "1 to 2 days",
    difftime(admin_time, procedure_date, units="days") > 2 & difftime(admin_time, procedure_date, units="days") <= 3 ~ "2 to 3 days",
    difftime(admin_time, procedure_date, units="days") > 3 & difftime(admin_time, procedure_date, units="days") <= 4 ~ "3 to 4 days",
  )
)

# Count antiemetic administrations per subject
total_admin <- ponv_medications %>%
  count(subject_id, name = "total_admin")

# Add column for admission duration (days) to ponv_study_cohort_features
ponv_study_cohort_features$admission_duration_days <- as.numeric(
  difftime(
    ponv_study_cohort_features$hosp_dischtime,
    ponv_study_cohort_features$hosp_admittime,
    units = "days"
  )
)

# Join onto the study cohort features
ponv_study_cohort_features <- ponv_study_cohort_features %>%
  left_join(total_admin, by = "subject_id") %>%
  mutate(total_admin = coalesce(total_admin, 0))

# Remove noise observations from ponv_study_cohort_features (if clustering method is HDBSCAN)
ponv_cluster_characteristics <- ponv_study_cohort_features %>% filter(cluster != 0)

# Count the total number of surgical patients in the final study cohort
sprintf("There are %s surgical patients in the final study cohort.", nrow(ponv_study_cohort_features))

# Count the number of noise observations for HDBSCAN 
if (method == "HDBSCAN"){
sprintf("HDBSCAN labeled %s observations as noise.", ponv_study_cohort_features %>% filter(cluster == 0) %>% count())
}
```

## Colors and Themes

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Colorblind friendly cluster colors 
cluster_colors <- c(
  "1" = "#56B4E9",
  "2" = "#E69F00",  
  "3" = "#009E73",  
  "4" = "#0072B2",  
  "5" = "#D55E00",  
  "6" = "#CC79A7",
  "7" = "#F0E442",
  "8" = "#999999")

# Nature themed ggplot
theme_nature <- theme_minimal(base_size = 11, base_family = "") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black", size = 0.3),
    axis.ticks = element_line(color = "black", size = 0.3),
    axis.text  = element_text(size = 11),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
    legend.position = "right",
    legend.text  = element_text(size = 11),
    legend.title = element_text(size = 11)
  )
```

# Cluster Characteristics

### Cluster Summary

```{r, echo=FALSE, message=FALSE}
# Cluster antiemetic medication use summary

# Cluster counts 
cluster_counts <- ponv_cluster_characteristics %>% count(cluster, antiemetics_use)
kable(cluster_counts, caption = "Cluster counts")

# Summary of subject-level antiemetic medication administrations by cluster
med_use_summary <- ponv_cluster_characteristics %>% group_by(cluster) %>% summarize(min(total_admin), median(total_admin), max(total_admin)) %>% ungroup()

kable(med_use_summary, caption = "Summary of subject-level anitemetic medication administrations by cluster")

cluster_counts
```

### Cluster Summary Table for Antiemetic Medication Use

```{r, echo=FALSE, message=FALSE, warning=FALSE}
cluster_summary <- ponv_study_cohort_features %>%
  mutate(
    cluster = factor(cluster, labels = sort(unique(cluster))),
    antiemetics_use = factor(antiemetics_use, levels = c("No", "Yes"))
  ) %>%
  group_by(cluster) %>%
  summarise(
    count = n(),

    # Binary antiemetics medication use at the cluster level
    antiemetics_use_binary = ifelse(
      any(antiemetics_use == "Yes"),
      "Yes",
      "No"
    ),

    # Median (min, max) total antiemetic administrations
    total_admin = paste0(
      sprintf("%.0f", median(total_admin, na.rm = TRUE)),
      " (",
      sprintf("%.0f, %.0f",
              min(total_admin, na.rm = TRUE),
              max(total_admin, na.rm = TRUE)
      ),
      ")"
    ))

# Add column for cluster descriptions
cluster_phenotypes <- c(
  "Frequent antiemetics use, mainly rescue",
  "No antiemetics use",
  "Predominantly late rescue",
  "Prophylaxis, day before surgery",
  "Prophylaxis/early rescue, day of surgery",
  "Prophylaxis heavy, some rescue",
  "Prophlaxis/early rescue",
  "Early rescue"
)

cluster_summary$cluster_phenotypes <- cluster_phenotypes

colnames(cluster_summary) <- c(
  "Cluster",
  "Count",
  "Antiemetics Use",
  "Median Antiemetics Admin (Min, Max)",
  "Phenotype"
)

# Ensure named vector for colors
cluster_color_map <- cluster_colors
names(cluster_color_map) <- levels(cluster_summary$Cluster)

cluster_table_kable <- cluster_summary %>%
  kable(
    format = "html",
    escape = FALSE,
    align = "lcccc"
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = NULL,
    font_size = 14
  ) %>%

  # Fix column widths (this enables wrapping)
  column_spec(1, width = "1.5cm") %>%   # Cluster
  column_spec(2, width = "1.5cm") %>%   # Count
  column_spec(3, width = "3.0cm") %>%   # Antiemetics use 
  column_spec(4, width = "4.0cm") %>%   # Median Antiemetics Admin
  column_spec(5, width = "7.5cm") %>%   # Cluster phenotypes 
  column_spec(
    1,
    bold = TRUE,
    background = cluster_color_map[cluster_summary$Cluster],
    color = "black"
  )

cluster_table_kable

# Save table as a PNG
save_kable(
  cluster_table_kable,
  file = "figures/main/02_PONV/ponv_cluster_summary_table_antiemetics.png",
  zoom = 2
)
```

### Demographic and Clinical Features 

```{r, echo=FALSE, message = FALSE}
age_summary <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    age_mean = mean(age, na.rm = TRUE),
    age_sd   = sd(age, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  mutate(
    age_mean_sd = sprintf("%.1f ± %.1f", age_mean, age_sd)
  ) %>%
  select(cluster, age_mean_sd) %>%
  pivot_wider(
    names_from  = cluster,
    values_from = age_mean_sd
  ) %>%
  rename_with(~ paste("Cluster", .x))

# Apfel risk factors by cluster
# Female gender, nonsmoker, postoperative opioids
female_pct <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    percentage = round(
      100 * mean(gender == "Female", na.rm = TRUE),
      2
    ),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = cluster,
    values_from = percentage,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x))

nonsmoker_pct <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    percentage = round(
      100 * mean(curr_or_prev_smoker == "No", na.rm = TRUE),
      2
    ),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = cluster,
    values_from = percentage,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x))

postop_opioids_pct <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    percentage = round(
      100 * mean(opioids_day_1_to_2 == "Yes" | opioids_day_2_to_4 == "Yes", na.rm = TRUE),
      2
    ),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = cluster,
    values_from = percentage,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x))

# Create a column in ponv_cluster_characteristics for modified 
# Apfel risk index (scale of 0 to 3)
# Add 1 point for Female gender, 1 point for Nonsmoker (never smoked), 
# and 1 point for postopreative opioids
ponv_cluster_characteristics <- ponv_cluster_characteristics %>% mutate(postoperative_opioids = case_when((opioids_day_1_to_2 == "Yes" | opioids_day_2_to_4 == "Yes") ~ "Yes",
                                                                    TRUE ~ "No"))

ponv_cluster_characteristics <- ponv_cluster_characteristics %>% group_by(subject_id) %>%
  mutate(modified_apfel_risk_score = sum(gender == "Female", curr_or_prev_smoker == "No", postoperative_opioids == "Yes")) %>% ungroup()

modified_apfel_risk_score_summary <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    modified_apfel_risk_score_mean = mean(modified_apfel_risk_score, na.rm = TRUE),
    modified_apfel_risk_score_sd   = sd(modified_apfel_risk_score, na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  mutate(
    modified_apfel_risk_score_mean_sd = sprintf("%.2f ± %.2f", modified_apfel_risk_score_mean, modified_apfel_risk_score_sd)
  ) %>%
  select(cluster, modified_apfel_risk_score_mean_sd) %>%
  pivot_wider(
    names_from  = cluster,
    values_from = modified_apfel_risk_score_mean_sd
  ) %>%
  rename_with(~ paste("Cluster", .x))

# Combine features into a single table 
demographics <- rbind(age_summary, female_pct, nonsmoker_pct, postop_opioids_pct) %>% mutate(Feature = c("Mean Age", "Female (%)", "Nonsmoker (%)", "Postoperative Opioids (%)")) %>% select(Feature, everything())

kable(demographics)
```

```{r, echo=FALSE, message=FALSE}
# Surgery type 
surgery_type_pct <- ponv_cluster_characteristics %>%
  group_by(cluster, surgery_type) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(
    cluster_total = sum(n),
    percentage = round(100 * n / cluster_total,2)
  ) %>%
  ungroup() %>%
  select(-n, -cluster_total) %>%
  pivot_wider(
    names_from = cluster,
    values_from = percentage,
    values_fill = 0
  ) %>%
  rename_with(
    ~ paste("Cluster", .x),
    -surgery_type
  )

kable(surgery_type_pct, caption = "Surgery type percentages by cluster")
```

### Cluster Summary Table for Clinical Risk Factors 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Create postoperative opioids + modified Apfel risk score
ponv_cluster_characteristics <- ponv_cluster_characteristics %>%
  mutate(
    postoperative_opioids = case_when(
      opioids_day_1_to_2 == "Yes" | opioids_day_2_to_4 == "Yes" ~ "Yes",
      TRUE ~ "No"
    )
  ) %>%
  group_by(subject_id) %>%
  mutate(
    modified_apfel_risk_score = sum(
      gender == "Female",
      curr_or_prev_smoker == "No",
      postoperative_opioids == "Yes"
    )
  ) %>%
  ungroup()

# Build one summary table with clusters as rows
risk_factor_summary <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    `Mean Age` = sprintf("%.1f (%.1f)",
                         mean(age, na.rm = TRUE),
                         sd(age, na.rm = TRUE)),
    
    `Female (%)` = round(
      100 * mean(gender == "Female", na.rm = TRUE), 2
    ),
    
    `Nonsmoker (%)` = round(
      100 * mean(curr_or_prev_smoker == "No", na.rm = TRUE), 2
    ),
    
    `Postoperative Opioids (%)` = round(
      100 * mean(postoperative_opioids == "Yes", na.rm = TRUE), 2
    ),
    
    `Mean Modified Apfel RI` = sprintf("%.2f ± %.2f",
                                       mean(modified_apfel_risk_score, na.rm = TRUE),
                                       sd(modified_apfel_risk_score, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  select(
    cluster,
    `Mean Age`,
    `Female (%)`,
    `Nonsmoker (%)`,
    `Postoperative Opioids (%)`
  )

colnames(risk_factor_summary) <- c(
  "Cluster",
  "Mean Age (SD)",
  "% Female",
  "% Nonsmoker",
  "% Postoperative Opioids"
)

# Ensure named vector for colors
cluster_color_map <- cluster_colors
names(cluster_color_map) <- levels(risk_factor_summary$Cluster)

cluster_table_kable <- risk_factor_summary %>%
  kable(
    format = "html",
    escape = FALSE,
    align = "lcccc"
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = NULL,
    font_size = 14
  ) %>%

  # Fix column widths (this enables wrapping)
  column_spec(1, width = "1.5cm") %>%   # Cluster
  column_spec(2, width = "3.0cm") %>%   # Age
  column_spec(3, width = "3.0cm") %>%   # Percent female
  column_spec(4, width = "3.5cm") %>%   # Nonsmoker percentage
  column_spec(5, width = "3.5cm") %>%   # Postoperative opioids 
  column_spec(
    1,
    bold = TRUE,
    background = cluster_color_map[risk_factor_summary$Cluster],
    color = "black"
  )

cluster_table_kable

# Save table as a PNG
save_kable(
  cluster_table_kable,
  file = "figures/main/02_PONV/ponv_cluster_summary_table_clinical_risk_factors.png",
  zoom = 2
)
```

### Modified Apfel Risk Score

The modified Apfel risk score is calculated on a scale of 0-3. We add 1 point for the following categories: female gender, nonsmoker, postoperative opiods (i.e., received opioids 1 to 4 days after surgery).

The normal Apfel risk index for PONV is on a scale of 0-4 and also includes a point for history of PONV or motion sickness. However, this information is unavailable in MIMIC-IV. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Summarize mean + SE modified Apfel RI by cluster 
summary_df <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    n = n(),
    mean_apfel = mean(modified_apfel_risk_score, na.rm = TRUE),
    sd_apfel   = sd(modified_apfel_risk_score, na.rm = TRUE),
    se_apfel   = sd_apfel / sqrt(n)
  )

# Bar plot displaying mean modified Apfel score with SE bars
modified_apfel_score_bar_plot <- ggplot() +
  
  geom_bar(
    data = summary_df,
    aes(x = factor(cluster), y = mean_apfel, fill = factor(cluster)),
    stat = "identity", width = 0.6, alpha = 1, color = "black"
  ) +
  
  scale_fill_manual(values = cluster_colors) +
  scale_color_manual(values = cluster_colors) +
  
  # Two-sided SE error bar
  geom_errorbar(
    data = summary_df,
    aes(
      x = factor(cluster),
      ymin = mean_apfel - se_apfel,
      ymax = mean_apfel + se_apfel
    ),
    width = 0.12,
    linewidth = 0.5
  ) +
  
  # mean labels ABOVE the SE bar
  geom_text(
    data = summary_df,
    aes(
      x = factor(cluster),
      y = mean_apfel + se_apfel + 0.15,
      label = round(mean_apfel, 2)
    ),
    size = 4.5,
    fontface = "bold"
  ) +
  
  labs(
    x = "Cluster",
    y = "Modified Apfel risk score"
  ) +
  theme_nature +
  theme(legend.position = "none")

modified_apfel_score_bar_plot

ggsave("figures/main/02_PONV/ponv_modified_apfel_scores_bar_plot.png",
  modified_apfel_score_bar_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```

## Cluster Visualization

### 2D PCA Plot

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# 2D PCA visualization of the clusters

# PCA
embedding_mat <- embeddings %>%
  arrange(subject_id) %>%
  select(-subject_id)

pca_result <- prcomp(embedding_mat, center = TRUE, scale. = FALSE)

pca_data <- as.data.frame(pca_result$x[, 1:2]) %>%
  mutate(subject_id = embeddings %>% arrange(subject_id) %>% pull(subject_id))

# Join cluster labels safely
plot_df <- pca_data %>%
  left_join(
    ponv_cluster_characteristics %>%
      select(subject_id, cluster),
    by = "subject_id"
  ) %>%
  mutate(Cluster = factor(cluster))

# PCA Plot
pca_plot <- ggplot(plot_df, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = cluster_colors) +
  labs(
    x = "PC 1",
    y = "PC 2"
  ) +
  theme_nature

pca_plot

ggsave("figures/main/02_PONV/ponv_2d_pca_plot.png",
  pca_plot,
  width  = 6.5,
  height = 4.5,
  dpi    = 300,
  bg     = "white")
```
### PCA Insights 

```{r, echo=TRUE, message=FALSE}
# Check the proportion of variation explained by the first 10 principle components 
pve <- round(pca_result$sdev^2 / sum(pca_result$sdev^2), 3)
pve[1:10]

# Check the correlation between the first 10 principle components and the total number of medication administrations 
cor(pca_result$x[, 1:10], ponv_cluster_characteristics$total_admin)
# There is a strong positive correlation between PC1 and the total number of medication administrations  

# Check the correlation between the first 10 principle components and age 
cor(pca_result$x[, 1:10], ponv_cluster_characteristics$age)

# Check the point-biserial correlation between the first 10 principle components and gender
gender_binary <- ifelse(
  ponv_cluster_characteristics$gender == "Male", 1, 0
)

pc_gender_cor <- apply(
  pca_result$x[, 1:10],
  2,
  function(pc) cor(pc, gender_binary, use = "complete.obs")
)

pc_gender_cor
```

### 2D UMAP Plot

```{r, eval=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(uwot)

# Prepare embedding matrix
embedding_df <- embeddings %>%
  arrange(subject_id)

embedding_mat <- embedding_df %>%
  select(-subject_id) %>%
  as.matrix()

# 2D UMAP
set.seed(42)

umap_result <- uwot::umap(
  embedding_mat,
  n_neighbors = 50,
  min_dist = 0.3,
  n_components = 2,
  metric = "cosine",
  scale = FALSE,
  init = "spectral",
  n_threads = parallel::detectCores() - 1,
  verbose = FALSE
)

# Convert to plotting data
umap_data <- as.data.frame(umap_result) %>%
  rename(UMAP1 = V1, UMAP2 = V2) %>%
  mutate(subject_id = embedding_df$subject_id)

plot_df <- umap_data %>%
  left_join(
    ponv_cluster_characteristics %>%
      select(subject_id, cluster),
    by = "subject_id"
  ) %>%
  mutate(Cluster = factor(cluster))

# UMAP plot
umap_plot <- ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = Cluster)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = cluster_colors) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2"
  ) +
  theme_nature

umap_plot

ggsave(
  "figures/main/02_PONV/ponv_2d_umap_plot.png",
  umap_plot,
  width = 6.5,
  height = 4.5,
  dpi = 300,
  bg = "white"
)
```

# Antiemetic Medications 

### Overall Medication Use

```{r, echo=FALSE, message=FALSE}
ggplot(
  ponv_medications %>%
    mutate(
      medication = fct_infreq(medication) |> fct_rev()
    ),
  aes(x = medication, fill = medication)
) +
  geom_bar(show.legend = FALSE) +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    hjust = -0.2,
    size = 3.5
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.25)))+
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Study cohort antiemetic medication counts",
    x = "Medication",
    y = "Number of administrations"
  )
```
```{r, echo=FALSE, message=FALSE}
ggplot(
  ponv_medications %>%
    mutate(
      drug_class = fct_infreq(drug_class) |> fct_rev()
    ),
  aes(x = drug_class, fill = drug_class)
) +
  geom_bar(show.legend = FALSE) +
  geom_text(
    stat = "count",
    aes(label = after_stat(count)),
    hjust = -0.2,
    size = 3.5
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.3)))+
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Study cohort antiemetic drug class counts",
    x = "Drug class",
    y = "Number of administrations"
  )
```

### Medication Types by Cluster

```{r, echo=FALSE, message=FALSE}
# Summarize the antiemetic medication administration percentages by cluster
medication_pct_table <- ponv_medications %>%
  filter(cluster != 0) %>%
  group_by(cluster, medication) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(
    percentage = round(100 * n / sum(n), 2)
  ) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(
    names_from = cluster,
    values_from = percentage,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x), -medication)

kable(medication_pct_table, caption = "Antiemetic Medication Administration Percentages by Cluster")
```

### Radar Plot for Medication Types 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Radar plot of medication types by cluster
medication_counts <- ponv_medications %>% 
  filter(cluster != 0) %>%
  group_by(cluster, medication) %>%
  summarise(n_admin = n(), .groups = "drop_last") %>%
  mutate(percentage = round(100 * n_admin / sum(n_admin),2)) %>%
  arrange(cluster, desc(percentage))

# Step 1: Wide format (clusters as rows, meds as columns)
radar_data <- medication_counts %>%
  select(cluster, medication, percentage) %>%
  pivot_wider(names_from = medication, values_from = percentage, values_fill = 0)

# radar_data <- radar_data %>%
#   select(cluster, all_of(medication_order))

# Step 2: Add required max/min rows for fmsb
radar_scaled <- rbind(
  rep(100, ncol(radar_data)-1),
  rep(0, ncol(radar_data)-1),
  radar_data[,-1]
)
rownames(radar_scaled) <- c("max", "min", paste0("Cluster ", radar_data$cluster))

# Extract colors in cluster order
radar_cluster_ids <- as.character(radar_data$cluster)
radar_colors <- cluster_colors[radar_cluster_ids]

# Save as a png image
png(
  filename = "figures/main/02_PONV/ponv_antiemetic_types_pct_radar.png",
  width  = 7.0,      # inches
  height = 4.5,      # inches
  units  = "in",
  res    = 300,      # DPI for journals
  bg     = "white"
)

# Adjust the margins 
par(mar = c(0, 0, 0.5, 1.3)) # bottom, left, top, right

fmsb::radarchart(radar_scaled,
                 axistype = 0,
                 pcol = radar_colors,
                 plwd = 0.5, # line width
                 plty = 1,
                 cglcol = "grey",
                 cglty = 1,
                 axislabcol = "grey20",
                 caxislabels = c(0, 20, 40, 60, 80, 100),
                 cglwd = 0.8,
                 vlcex = 0.8,
                 cex.main = 0.8)

legend("topright",
       legend = paste0("Cluster ", radar_cluster_ids),
       col = radar_colors,
       lty = 1, lwd = 2, cex = 0.8)

# Close device 
dev.off()
```

```{r, echo=FALSE, message=FALSE}
# Summarize subject-level antiemetic medication counts by cluster  

# ---- Complete medication list ----
all_medications <- unique(ponv_medications$medication)

# ---- All patients per cluster ----
cluster_patients <- ponv_medications %>%
  filter(cluster != 0) %>%
  distinct(cluster, subject_id)

# ---- Full patient × medication grid ----
patient_med_grid <- cluster_patients %>%
  crossing(medication = all_medications)

# ---- Observed administration counts ----
admin_counts <- ponv_medications %>%
  filter(cluster != 0) %>%
  group_by(cluster, subject_id, medication) %>%
  summarise(n_admin = n(), .groups = "drop")

# ---- Fill missing administrations with zeros ----
admin_counts_complete <- patient_med_grid %>%
  left_join(
    admin_counts,
    by = c("cluster", "subject_id", "medication")
  ) %>%
  mutate(n_admin = replace_na(n_admin, 0))

# ---- Median administrations per patient ----
medication_median_table <- admin_counts_complete %>%
  group_by(cluster, medication) %>%
  summarise(
    median_admin = median(n_admin),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = cluster,
    values_from = median_admin,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x), -medication)

# ---- Mean administrations per patient ----
medication_mean_table <- admin_counts_complete %>%
  group_by(cluster, medication) %>%
  summarise(
    mean_admin = round(mean(n_admin),2),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from  = cluster,
    values_from = mean_admin,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x), -medication)

# ---- Median administrations per patient among those exposed to each medication ----
mean_exposed <- ponv_medications %>%
  filter(cluster != 0) %>%
  group_by(cluster, subject_id, medication) %>%
  summarise(n_admin = n(), .groups = "drop") %>%
  group_by(cluster, medication) %>%
  summarise(
    mean_admin_exposed = round(mean(n_admin),2),
    .groups = "drop"
  ) %>%
   pivot_wider(
    names_from  = cluster,
    values_from = mean_admin_exposed,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x), -medication)

kable(medication_median_table, caption = "Median administrations per patient")

kable(medication_mean_table, caption = "Mean administrations per patient")

kable(medication_mean_table, caption = "Mean administrations per patient (among patients exposed to each medication)")
```

### Drug Class Percentages 

```{r, echo=FALSE, message=FALSE}
medication_pct_table <- ponv_medications %>%
  filter(cluster != 0) %>%
  group_by(cluster, drug_class) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(
    percentage = round(100 * n / sum(n), 2)
  ) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(
    names_from = cluster,
    values_from = percentage,
    values_fill = 0
  ) %>%
  rename_with(~ paste("Cluster", .x), -drug_class)

kable(medication_pct_table, caption = "Antiemetic Drug Class Percentages by Cluster")
```

### Radar Plot for Drug Classes

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Radar plot of drug class percentages by cluster

# Set the order of drug classes in the radar plot
drug_class_order <- c("5-HT3 receptor antagonists", "Benzamides", "Corticosteroids", "Antihistamines", "Phenothiazines", "Anticholinergics", "Butyrophenones", "Neurokinin-1 antagonists")

drug_class_counts <- ponv_medications %>% 
  filter(cluster != 0) %>%
  mutate(drug_class = factor(drug_class, levels = drug_class_order)) %>%
  group_by(cluster, drug_class) %>%
  summarise(n_admin = n(), .groups = "drop_last") %>%
  mutate(percentage = round(100 * n_admin / sum(n_admin),2)) %>%
  arrange(cluster, desc(percentage))

# Step 1: Wide format (clusters as rows, meds as columns)
radar_data <- drug_class_counts %>%
  select(cluster, drug_class, percentage) %>%
  pivot_wider(names_from = drug_class, values_from = percentage, values_fill = 0)

radar_data <- radar_data %>%
  select(cluster, all_of(drug_class_order))

# Step 2: Add required max/min rows for fmsb
radar_scaled <- rbind(
  rep(100, ncol(radar_data)-1),
  rep(0, ncol(radar_data)-1),
  radar_data[,-1]
)
rownames(radar_scaled) <- c("max", "min", paste0("Cluster ", radar_data$cluster))

# Extract colors in cluster order
radar_cluster_ids <- as.character(radar_data$cluster)
radar_colors <- cluster_colors[radar_cluster_ids]

# Save as a png image
png(
  filename = "figures/main/02_PONV/ponv_antiemetic_drug_class_pct_radar.png",
  width  = 7.0,      # inches
  height = 4.5,      # inches
  units  = "in",
  res    = 300,      # DPI for journals
  bg     = "white"
)

# Adjust the margins 
par(mar = c(0, 0, 0.5, 1.3)) # bottom, left, top, right

fmsb::radarchart(radar_scaled,
                 axistype = 0,
                 pcol = radar_colors,
                 plwd = 0.5, # line width
                 plty = 1,
                 cglcol = "grey",
                 cglty = 1,
                 axislabcol = "grey20",
                 caxislabels = c(0, 20, 40, 60, 80, 100),
                 cglwd = 0.8,
                 vlcex = 0.8,
                 cex.main = 0.8)

legend("topright",
       legend = paste0("Cluster ", radar_cluster_ids),
       col = radar_colors,
       lty = 1, lwd = 2, cex = 0.75)

# Close device 
dev.off()
```

### Radar Plot for Administration Modes

```{r, echo=FALSE, message=FALSE, warning=FALSE}
route_counts <- ponv_medications %>% 
  filter(cluster != 0) %>%
  group_by(cluster, route) %>%
  summarise(n_admin = n(), .groups = "drop_last") %>%
  mutate(percentage = round(100 * n_admin / sum(n_admin),2)) %>%
  arrange(cluster, desc(percentage))

# Step 1: Wide format (clusters as rows, meds as columns)
radar_data <- route_counts %>%
  select(cluster, route, percentage) %>%
  pivot_wider(names_from = route, values_from = percentage, values_fill = 0)

# Step 2: Add required max/min rows for fmsb
radar_scaled <- rbind(
  rep(100, ncol(radar_data)-1),
  rep(0, ncol(radar_data)-1),
  radar_data[,-1]
)
rownames(radar_scaled) <- c("max", "min", paste0("Cluster ", radar_data$cluster))

# Extract colors in cluster order
radar_cluster_ids <- as.character(radar_data$cluster)
radar_colors <- cluster_colors[radar_cluster_ids]

# Save as a png image
png(
  filename = "figures/main/02_PONV/ponv_antiemetic_routes_radar.png",
  width  = 7.0,      # inches
  height = 4.5,      # inches
  units  = "in",
  res    = 300,      # DPI for journals
  bg     = "white"
)

# Adjust the margins 
par(mar = c(0, 0, 0, 2.8)) # bottom, left, top, right

fmsb::radarchart(radar_scaled,
                 axistype = 0,
                 pcol = radar_colors,
                 plwd = 0.5, # line width
                 plty = 1,
                 cglcol = "grey",
                 cglty = 1,
                 axislabcol = "grey20",
                 caxislabels = c(0, 20, 40, 60, 80, 100),
                 cglwd = 0.8,
                 vlcex = 0.9, # route label sizes
                 cex.main = 0.8)

legend("topright",
       legend = paste0("Cluster ", radar_cluster_ids),
       col = radar_colors,
       lty = 1, lwd = 1.4, cex = 0.8,
       inset = c(-0.05, 0.05),
       xpd = TRUE)

# Close device
dev.off()
```

### Administration Timing (Percentage receiving antiemetics)

```{r, echo=FALSE}
antiemetics_day_before_surgery_pct <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    day_before_surgery_pct = round(
      100 * mean(antiemetics_day_before_surgery == "Yes", na.rm = TRUE),
      2
    ),
    .groups = "drop"
  )

antiemetics_day_0_to_1_pct <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    day_0_to_1_pct = round(
      100 * mean(antiemetics_day_0_to_1 == "Yes", na.rm = TRUE),
      2
    ),
    .groups = "drop"
  )

antiemetics_day_1_to_2_pct <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    day_1_to_2_pct = round(
      100 * mean(antiemetics_day_1_to_2 == "Yes", na.rm = TRUE),
      2
    ),
    .groups = "drop"
  )

antiemetics_day_2_to_4_pct <- ponv_cluster_characteristics %>%
  group_by(cluster) %>%
  summarise(
    day_2_to_4_pct = round(
      100 * mean(antiemetics_day_2_to_4 == "Yes", na.rm = TRUE),
      2
    ),
    .groups = "drop"
  )

all_time_periods <- cbind(antiemetics_day_before_surgery_pct, antiemetics_day_0_to_1_pct %>% select(-cluster), antiemetics_day_1_to_2_pct %>% select(-cluster), antiemetics_day_2_to_4_pct %>% select(-cluster))

all_time_periods_long <- all_time_periods %>%
  pivot_longer(
    cols = c(
      day_before_surgery_pct,
      day_0_to_1_pct,
      day_1_to_2_pct,
      day_2_to_4_pct
    ),
    names_to  = "time_window",
    values_to = "percentage"
  ) %>%
   mutate(
    time_window = factor(
      recode(
        time_window,
        day_before_surgery_pct = "Day before surgery",
        day_0_to_1_pct         = "Day 0–1",
        day_1_to_2_pct         = "Day 1–2",
        day_2_to_4_pct         = "Day 2–4"
      ),
      levels = c(
        "Day before surgery",
        "Day 0–1",
        "Day 1–2",
        "Day 2–4"
      )
    )
  )

time_window_plot <- ggplot(
  all_time_periods_long,
  aes(x = factor(cluster), y = percentage, fill = factor(cluster))
) +
  geom_col() +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  facet_wrap(~ time_window, nrow = 1) +
  labs(
    x = "Cluster",
    y = "Percentage of patients receiving antiemetics"
  ) +
  theme_nature

time_window_plot
```


### Administration Timing (Total counts)

```{r, echo=FALSE}
all_time_periods_long <- ponv_medications %>%
  filter(cluster != 0) %>%
  filter(admin_interval %in% c("-1 to 0 days", "0 to 1 days", "1 to 2 days", "2 to 3 days", "3 to 4 days")) %>%
  count(cluster, admin_interval, name = "n_admin") %>%
  tidyr::complete(
    cluster = 1:8,
    admin_interval = c("-1 to 0 days", "0 to 1 days", "1 to 2 days", "2 to 3 days", "3 to 4 days"),
    fill = list(n_admin = 0)
  ) %>%
  mutate(
    time_window = factor(
      recode(
        admin_interval,
        "-1 to 0 days" = "Day before surgery",
        "0 to 1 days"  = "Day 0–1",
        "1 to 2 days"  = "Day 1–2",
        "2 to 3 days"  = "Day 2–3",
        "3 to 4 days"  = "Day 3–4"
      ),
      levels = c("Day before surgery", "Day 0–1", "Day 1–2", "Day 2–3", "Day 3–4")
    ),
    cluster = factor(cluster, levels = 1:8)
  )

time_window_plot <- ggplot(
  all_time_periods_long,
  aes(x = cluster, y = n_admin, fill = cluster)
) +
  geom_col() +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  facet_wrap(~ time_window, nrow = 1) +
  labs(
    x = "Cluster",
    y = "Total administration count"
  ) +
  theme_nature

time_window_plot
```

### Administration Timing (Counts per patient) 

```{r, echo=FALSE, message=FALSE}
# Number of patients in each cluster
cluster_sizes <- ponv_medications %>%
  filter(cluster != 0) %>%
  distinct(subject_id, cluster) %>%
  count(cluster, name = "n_patients")

# Administration counts per cluster and time window
all_time_periods_long <- ponv_medications %>%
  filter(cluster != 0) %>%
  filter(admin_interval %in% c("-1 to 0 days", "0 to 1 days", "1 to 2 days", "2 to 3 days", "3 to 4 days")) %>%
  count(cluster, admin_interval, name = "n_admin") %>%
  tidyr::complete(
    cluster = 1:8,
    admin_interval = c("-1 to 0 days", "0 to 1 days", "1 to 2 days", "2 to 3 days", "3 to 4 days"),
    fill = list(n_admin = 0)
  ) %>%
  left_join(cluster_sizes, by = "cluster") %>%
  mutate(
    admin_per_patient = n_admin / n_patients,
    time_window = factor(
      recode(
        admin_interval,
        "-1 to 0 days" = "Day before surgery",
        "0 to 1 days"  = "Day 0–1",
        "1 to 2 days"  = "Day 1–2",
        "2 to 3 days"  = "Day 2–3",
        "3 to 4 days"  = "Day 3–4"
      ),
      levels = c("Day before surgery", "Day 0–1", "Day 1–2", "Day 2–3", "Day 3–4")
    ),
    cluster = factor(cluster, levels = 1:8)
  )

# Max y value
y_max <- max(all_time_periods_long$admin_per_patient, na.rm = TRUE)

# Only ONE segment for the first facet (far left)
axis_segment <- data.frame(
  time_window = factor("Day before surgery",
    levels = levels(all_time_periods_long$time_window)
  ),
  x = 0.5,
  xend = 0.5,
  y = -0.04 * y_max,          # slightly below 0 to connect to x-axis
  yend = y_max * 1.15         # slightly below top to create gap
)

admins_per_patient_over_time_plot <- ggplot(
  all_time_periods_long,
  aes(x = cluster, y = admin_per_patient, fill = cluster)
) +
  geom_col() +
  geom_segment(
    data = axis_segment,
    aes(x = x, xend = xend, y = y, yend = yend),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 0.3
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(
  expand = expansion(mult = c(0, 0.08))
)+
  facet_wrap(~ time_window, nrow = 1) +
  labs(
    x = "Cluster",
    y = "Administrations per patient"
  ) +
  theme_nature +
  theme(
    axis.line.y = element_blank(),   # remove default y-axis everywhere
    strip.text = element_text(size = 8),
    strip.background = element_rect(fill = "grey95", color = "black", linewidth = 0.3),
    panel.spacing = unit(1.05, "lines"),
    axis.text.x = element_text(size = 10),
    axis.title = element_text(size = 10),
    panel.grid.major.y = element_line(color = "grey88", linewidth = 0.25)
  )

admins_per_patient_over_time_plot

ggsave(
  "figures/main/02_PONV/ponv_antiemetic_admins_per_patient_over_time.png",
  admins_per_patient_over_time_plot,
  width  = 7.3,
  height = 4.0,
  dpi    = 300,
  bg     = "white")
```

### Distinct Drug Classes Summary 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ponv_medications %>%
  group_by(cluster, subject_id) %>%
  summarise(
    n_distinct_drug_classes = n_distinct(drug_class),
    .groups = "drop"
  ) %>%
  group_by(cluster) %>%
  summarise(
    avg_distinct_drug_classes = mean(n_distinct_drug_classes),
    sd_distinct_drug_classes  = sd(n_distinct_drug_classes),
    n_patients = n(),
    .groups = "drop"
  )
```
### Drug class switches (percentage switching per cluster)

```{r, echo=FALSE, message=FALSE}
# Order medications and identify switches in drug class within each subject
ponv_switches <- ponv_medications %>%
  arrange(subject_id, admin_time) %>%
  group_by(subject_id) %>%
  mutate(
    prev_class = lag(drug_class),
    is_switch = !is.na(prev_class) & drug_class != prev_class
  ) %>%
  ungroup()

# Subject-level indicator: any drug class switch
switch_subject <- ponv_switches %>%
  group_by(subject_id, cluster) %>%
  summarise(
    any_switch = any(is_switch),
    .groups = "drop"
  )

# Include all expected clusters, even if absent from switch_subject
all_clusters <- 1:8

switch_percent <- switch_subject %>%
  group_by(cluster) %>%
  summarise(
    n_switched       = sum(any_switch),
    n_subjects       = n(),
    percent_switched = 100 * n_switched / n_subjects,
    .groups = "drop"
  ) %>%
  complete(
    cluster = all_clusters,
    fill = list(
      n_switched = 0,
      n_subjects = 0,
      percent_switched = 0
    )
  )

# Plot
switch_pct_plot <- ggplot(
  switch_percent,
  aes(x = factor(cluster), y = percent_switched, fill = factor(cluster))
) +
  geom_col(width = 0.7) +
  geom_text(
    aes(
      label = paste0(
        sprintf("%.1f%%", percent_switched),
        "\n(n=", n_switched, ")"
      )
    ),
    vjust = -0.3,
    size = 3.5
  ) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  labs(
    x = "Cluster",
    y = "Patients with ≥1 drug class switch (%)"
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_nature

switch_pct_plot

ggsave(
  "figures/main/02_PONV/ponv_drug_class_switch_percentages_by_cluster.png",
  switch_pct_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

### Drug class switches (switch counts by cluster)

```{r, echo=FALSE, message=FALSE}
switch_events <- ponv_medications %>%
  arrange(cluster, subject_id, admin_time) %>%
  group_by(cluster, subject_id) %>%
  mutate(
    prev_class = lag(drug_class),
    prev_time  = lag(admin_time),
    time_diff_hours = as.numeric(difftime(admin_time, prev_time, units = "hours"))
  ) %>%
  ungroup() %>%
  filter(
    !is.na(prev_class),
    prev_class != drug_class
  )

switch_counts <- switch_events %>%
  count(cluster, prev_class, drug_class, name = "n") %>%
  group_by(cluster) %>%
  mutate(
    pct = 100 * n / sum(n)
  ) %>%
  ungroup() %>%
  left_join(
    switch_events %>%
      group_by(cluster, prev_class, drug_class) %>%
      summarise(
        mean_time_hours = mean(time_diff_hours, na.rm = TRUE),
        median_time_hours = median(time_diff_hours, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("cluster", "prev_class", "drug_class")
  ) %>%
  group_by(cluster) %>%
  arrange(desc(pct), .by_group = TRUE) %>%
  ungroup()
```

### Most Common Drug Class Switches Table

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Make sure cluster is treated consistently
switch_table_data <- switch_counts %>%
  mutate(
    cluster = as.character(cluster),
    switch = paste(prev_class, "\u2192", drug_class),
    pct = round(pct, 1)
  ) %>%
  filter(cluster %in% c("1", "3", "6", "7")) %>%
  group_by(cluster) %>%
  arrange(desc(pct), .by_group = TRUE) %>%
  slice_head(n = 2) %>%
  mutate(
    rank = row_number(),
    switch_label = paste0(switch, " (", pct, "%)")
  ) %>%
  ungroup() %>%
  select(cluster, rank, switch_label) %>%
  pivot_wider(
    names_from = rank,
    values_from = switch_label,
    names_prefix = "Top "
  ) %>%
  rename(
    Cluster = cluster,
    `Most Common Switch` = `Top 1`,
    `Second Most Common Switch` = `Top 2`
  )

# Ensure named vector for colors
cluster_color_map <- cluster_colors
names(cluster_color_map) <- as.character(sort(unique(ponv_study_cohort_features$cluster)))

switch_table_kable <- switch_table_data %>%
  kable(
    format = "html",
    escape = FALSE,
    align = "lcc",
    col.names = c("Cluster", "Most Common Switch", "Second Most Common Switch")
  ) %>%
  kable_styling(
    full_width = FALSE,
    bootstrap_options = NULL,
    font_size = 14
  ) %>%
  column_spec(1, width = "1.5cm", bold = TRUE,
              background = cluster_color_map[switch_table_data$Cluster],
              color = "black") %>%
  column_spec(2, width = "6.5cm") %>%
  column_spec(3, width = "6.5cm")

switch_table_kable

# Save table as PNG
save_kable(
  switch_table_kable,
  file = "figures/main/02_PONV/ponv_most_common_switches_table.png")
```

# Health Outcomes 

### Discharge Location 

```{r, echo=FALSE, message=FALSE}
# Group discharge locations into common categories 
ponv_discharge_location <- ponv_study_cohort_features %>%
  mutate(
    discharge_location = case_when(
      discharge_location %in% c("HOME", "HOME HEALTH CARE") ~ "Home/HHC",
      discharge_location == "REHAB" ~ "Rehab",
      discharge_location == "SKILLED NURSING FACILITY" ~ "SNF",
      discharge_location == "CHRONIC/LONG TERM ACUTE CARE" ~ "LTACH",
      discharge_location %in% c(
        "PSYCH FACILITY", "OTHER FACILITY", "ASSISTED LIVING",
        "AGAINST ADVICE", "ACUTE HOSPITAL", "HEALTHCARE FACILITY", 
        "Unknown"
      ) ~ "Other",
      discharge_location == "HOSPICE" ~ "Hospice",
      discharge_location == "DIED" ~ "Died"
    ),
    discharge_location = factor(
      discharge_location,
      levels = c(
        "Died",
        "Hospice",
        "Other",
        "LTACH",
        "SNF",
        "Rehab",
        "Home/HHC"
      ),
      ordered = TRUE
    )
  )

discharge_pct <- ponv_discharge_location %>%
  mutate(
    cluster = factor(cluster),
    discharge_location = factor(discharge_location)
  ) %>%
  group_by(cluster, discharge_location) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(cluster) %>%
  mutate(
    pct = 100 * n / sum(n)
  )

discharge_locations_plot <- ggplot(
  discharge_pct,
  aes(x = discharge_location, y = pct, fill = cluster)
) +
  geom_col(width = 0.8) +
  facet_wrap(
  ~ cluster,
  ncol = 4,
  labeller = labeller(cluster = function(x) paste("Cluster", x))
) +
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(
    #labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    x = "Discharge location",
    y = "Percentage of patients"
  ) +
  geom_text(
  aes(
    label = paste0(
      sprintf("%.1f%%", pct)),
    y = pct
  ),
  hjust = -0.05,
  size = 2.4
) +
  coord_flip() + theme_nature

discharge_locations_plot 

ggsave(
  "figures/main/02_PONV/ponv_discharge_locations_bar_plot.png",
  discharge_locations_plot,
  width  = 9,
  height = 4,
  dpi    = 300,
  bg     = "white")
```

### Hospital Admission Duration Bar Plot

```{r, echo=FALSE, message=FALSE}
admission_duration_summary <- ponv_study_cohort_features %>%
  group_by(cluster) %>%
  summarise(
    mean_admission_duration_days = mean(admission_duration_days),
    sd_admission_duration_days   = sd(admission_duration_days),
    n_patients      = n(),
    .groups = "drop"
  )

admission_duration_plot <- ggplot(
  admission_duration_summary,
  aes(
    x = factor(cluster),
    y = mean_admission_duration_days,
    fill = factor(cluster)
  )
) +
  geom_col(width = 0.7) +
  geom_errorbar(
    aes(
      ymin = mean_admission_duration_days - sd_admission_duration_days / sqrt(n_patients),
      ymax = mean_admission_duration_days + sd_admission_duration_days / sqrt(n_patients)
    ),
    width = 0.2
  ) +
geom_text(
  aes(
    y = mean_admission_duration_days + sd_admission_duration_days / sqrt(n_patients),
    label = sprintf("%.1f", mean_admission_duration_days)
  ),
  vjust = -0.7,
  size = 3.5
)+
  scale_fill_manual(values = cluster_colors, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "Cluster",
    y = "Mean hospital admission duration (days)",
  ) + theme_nature 

admission_duration_plot

ggsave(
  "figures/main/02_PONV/ponv_admission_duration_bar_plot.png",
  admission_duration_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)
```

### Hospital Admission Duration Violin Plot

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Compute medians
admission_duration_medians <- ponv_study_cohort_features %>%
  group_by(cluster) %>%
  summarise(
    median_admission_duration_days = median(admission_duration_days, na.rm = TRUE),
    .groups = "drop"
  )

# Plot
admission_duration_plot <- ggplot(
  ponv_study_cohort_features,
  aes(
    x = factor(cluster),
    y = admission_duration_days,
    fill = factor(cluster)
  )
) +
  geom_violin(trim = FALSE, alpha = 0.9) +

  # Median point
  geom_point(
    data = admission_duration_medians,
    aes(
      x = factor(cluster),
      y = median_admission_duration_days
    ),
    shape = 21,
    size = 2.5,
    fill = "white"
  ) +

  # Median label
  geom_text(
    data = admission_duration_medians,
    aes(
      x = factor(cluster),
      y = median_admission_duration_days,
      label = sprintf("%.1f", median_admission_duration_days)
    ),
    vjust = -0.8,
    size = 3.5
  ) +

  scale_fill_manual(values = cluster_colors, guide = "none") +

  scale_y_log10(
    expand = expansion(mult = c(0, 0.15))
  ) +

  labs(
    x = "Cluster",
    y = "Hospital admission duration (days, log scale)"
  ) +
  theme_nature

ggsave(
  "figures/main/02_PONV/ponv_admission_duration_violin_plot.png",
  admission_duration_plot,
  width  = 7.0,
  height = 4.5,
  dpi    = 300,
  bg     = "white"
)

admission_duration_plot
```

# Hypothesis Testing 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Perform hypothesis tests to compare subgroups with Benjamini-Hochberg FDR adjusted p-values

# Ensure that cluster is a factor variable
ponv_cluster_characteristics$cluster <- factor(ponv_cluster_characteristics$cluster)

# Modified Apfel risk score (Kruskal–Wallis)
ponv_cluster_characteristics <- ponv_cluster_characteristics %>%
  mutate(
    postoperative_opioids = case_when(
      opioids_day_1_to_2 == "Yes" | opioids_day_2_to_4 == "Yes" ~ "Yes",
      TRUE ~ "No"
    )
  ) %>%
  group_by(subject_id) %>%
  mutate(
    modified_apfel_risk_score = sum(
      gender == "Female",
      curr_or_prev_smoker == "No",
      postoperative_opioids == "Yes"
    )
  ) %>%
  ungroup()

apfel_kw <- kruskal.test(modified_apfel_risk_score ~ cluster, data = ponv_cluster_characteristics)

p_apfel_kw <- apfel_kw$p.value

# Hospital length of stay (Kruskal-Wallis)
hospital_los_kw <- kruskal.test(admission_duration_days ~ cluster, data = ponv_cluster_characteristics)

p_hospital_los_kw <- hospital_los_kw$p.value

# Discharge location (Chi-square)
ponv_discharge_location <- ponv_cluster_characteristics %>%
  mutate(
    discharge_location = case_when(
      discharge_location %in% c("HOME", "HOME HEALTH CARE") ~ "Home/HHC",
      discharge_location == "REHAB" ~ "Rehab",
      discharge_location == "SKILLED NURSING FACILITY" ~ "SNF",
      discharge_location == "CHRONIC/LONG TERM ACUTE CARE" ~ "LTACH",
      discharge_location %in% c(
        "PSYCH FACILITY", "OTHER FACILITY", "ASSISTED LIVING",
        "AGAINST ADVICE", "ACUTE HOSPITAL", "HEALTHCARE FACILITY",
        "Unknown"
      ) ~ "Other",
      discharge_location == "HOSPICE" ~ "Hospice",
      discharge_location == "DIED" ~ "Died"
    ),
    discharge_location = factor(
      discharge_location,
      levels = c(
        "Died",
        "Hospice",
        "Other",
        "LTACH",
        "SNF",
        "Rehab",
        "Home/HHC"
      ),
      ordered = FALSE
    )
  )

discharge_chisq <- chisq.test(table(ponv_discharge_location$discharge_location, ponv_discharge_location$cluster))

p_discharge_chisq <- discharge_chisq$p.value

# Calculate the BH adjusted p-values and display in table
# Format p-values
format_pval <- function(p) {
  ifelse(
    p < 0.001,
    "<0.001",
    sprintf("%.3f", p)
  )
}

# Calculate BH-adjusted p-values
test <- c(
  "Modified Apfel risk score (Kruskal–Wallis)",
  "Hospital length of stay (Kruskal–Wallis)",
  "Discharge location (Chi-square)"
)

raw_pval <- c(
  p_apfel_kw, 
  p_hospital_los_kw,
  p_discharge_chisq)

p_adj <- p.adjust(raw_pval, method = "BH")

hypothesis_test_results <- data.frame(
  `Hypothesis test` = test,
  `Raw p-value` = format_pval(raw_pval),
  `BH-adjusted p-value` = format_pval(p_adj),
  check.names = FALSE
)

hypothesis_test_results <- hypothesis_test_results %>%
  mutate(`Hypothesis test` = str_replace_all(`Hypothesis test`, "Kruskal.{1,5}Wallis", "Kruskal-Wallis"))

kable(
  caption = "PONV Application Hypothesis Testing Results",
  hypothesis_test_results,
  align = c("l", "c", "c")
)

# Create LaTeX table
ponv_hypothesis_test_results_latex <- kable(
  hypothesis_test_results,
  format = "latex",
  booktabs = TRUE,
  longtable = FALSE,
  caption = "Hypothesis testing results for the PONV application.",
  label = "tab:ponv_hypothesis_tests",
  align = c("l", "c", "c"),
  escape = TRUE,
  linesep = ""
) %>%
  kable_styling(
    latex_options = "hold_position",
    position = "center"
  ) %>%
  column_spec(1, width = "9cm") %>%
  column_spec(2, width = "2.5cm") %>%
  column_spec(3, width = "3cm")

# Save table as a .tex file
dir.create("tables/02_PONV", recursive = TRUE, showWarnings = FALSE)
save_kable(ponv_hypothesis_test_results_latex,  "tables/02_PONV/ponv_hypothesis_test_results.tex")
```
